---
title: "250329_CrypticPSIfill"
author: "Irika Sinha"
date: "2025-02-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Purpose

The purpose of this script is to take the file with manually identified cryptic exon coordinates and use those coordinates to identify left and right junction PSIs in the different conditions for i3N. 

## Packages

Load packages

```{r packages}
CRANpackages <- c("tidyverse", "ggplot2", "ggpubr", "ggstats","BiocManager","viridis","pheatmap","readxl", "data.table")

for(i in CRANpackages){
  if(!(i %in% rownames(installed.packages()))){
    install.packages(pkgs = i)
  }
  library(package = i, character.only = T, quietly = T)
  rm(i)
}

#library("clusterProfiler","org.Hs.eg.db")

```

## Create a dataframe of splice junction counts
.sj files generated from STAR alignment of reads and include splice junction information for each sample.
Create a new dataframe with splice junction counts from each sample for relevant junctions

```{r mergeSJ, eval=FALSE}
### Load cryptic files
cryptic_df <- readxl::read_xlsx("/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Papers/NMD/02_Revisions/TDP43-NMD_Tables_Main.xlsx", sheet = 2)[,1:19]
skiptic_df <- readxl::read_xlsx("/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Papers/NMD/02_Revisions/TDP43-NMD_Tables_Main.xlsx", sheet = 3)[,1:5]

## Clean up junctions - cryptic_df
for(i in c(5:6,18:19)){
  cryptic_df[,i] = apply(cryptic_df[,i],2,function(x) gsub(pattern = ",", replacement = "", x = x))
  cryptic_df[,i] = apply(cryptic_df[,i],2,function(x) gsub(pattern = " ", replacement = "", x = x))
  rm(i)
}

## Clean up junctions - skiptic_df
for(i in c(3:5)){
  skiptic_df[,i] = apply(skiptic_df[,i],2,function(x) gsub(pattern = ",", replacement = "", x = x))
  skiptic_df[,i] = apply(skiptic_df[,i],2,function(x) gsub(pattern = " ", replacement = "", x = x))
  rm(i)
}

## Create a list of unique junctions being queried
uniqueJunctions = unique(c(cryptic_df$canonical_splice_junction, 
                         cryptic_df$left_junction, cryptic_df$right_junction,
                         skiptic_df$left_canonical_junction, skiptic_df$right_canonical_junction,
                         skiptic_df$skiptic_junction))
uniqueJunctions = unlist(as.list(str_split(uniqueJunctions,";")))

## Create empty dataframe to populate with splice junction counts
finaldf <- data.frame()

## Load .sj files and merge into one large data frame
# Filter to splice junctions being queried
sj_wd <- "/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Scripts/250106_CrypticCoordinateFill/SJfiles/"

for(i in list.files(sj_wd)){
  condName = gsub(pattern=".sj",replacement="",x = i)
  tempdf <- fread(paste0(sj_wd,i)) %>% 
    mutate(Condition = condName,
           Coordinates = paste0(V1,":",V2,"-",V3)) %>%
    filter(Coordinates %in% uniqueJunctions)
  
  finaldf <- rbind(finaldf,tempdf)
  rm(i,condName,tempdf)
}

## Isolate needed columns & Rename appropriately
finaldf <- cbind(finaldf, apply(tibble(finaldf$Condition),MARGIN = 2, 
                          function(x) str_split_i(string = x, pattern = "[:digit:]",i=1)))
finaldf <- finaldf %>% dplyr::select(!Condition)
finaldf <- finaldf[,4:11] %>% relocate(Coordinates, .before=1)
colnames(finaldf)[2:8] <- c("Strand","IntronMotif","Annotated","UniqueMap","MultiMap","MaxOverhang","Condition")

## Redefine strand and intron motif as per reference
finaldf <- finaldf %>% dplyr::select(!Annotated) %>% 
  mutate(Strand = ifelse(Strand == 1, "+", ifelse(Strand == 2, "-","Undef"))) %>%
  mutate(IntronMotif = ifelse(IntronMotif == 0, "noncanonical",
                              ifelse(IntronMotif ==1, "GT/AG",
                                     ifelse(IntronMotif == 2, "CT/AC",
                                            ifelse(IntronMotif == 3, "GC/AG",
                                                   ifelse(IntronMotif == 4, "CT/GC",
                                                          ifelse(IntronMotif == 5, "AT/AC",
                                                                 ifelse(IntronMotif == 6, "GT/AT", "noncanonical"))))))))
## Reframe dataframe
finaldf <- finaldf %>% filter(Condition != "FALSE") %>% 
  pivot_wider(names_from = "Condition", # Each column is a treatment
              values_from = "UniqueMap", # Values are the # of splice junction counts
              values_fn = function(x) mean(x, na.rm=T),  # Average splice junction counts between duplicates
              id_cols = "Coordinates")

# write.csv(finaldf,"output_tables/251224_SpliceJunctions_Cleaned.csv",row.names = F)
# write.csv(uniqueJunctions[which(!(uniqueJunctions %in% finaldf$Coordinates))],"output_tables/251224_TestJunctions.csv")
```

## Calculate the PSI



```{r calcPSI, eval=FALSE}
junctionCountdf <- fread("output_tables/251224_SpliceJunctions_Cleaned.csv")

## Create new dataframe to populate with splice junction counts: cryptic exons
cryptic_df_fill = cryptic_df
cryptic_df_fill[,20:30] = NA
colnames(cryptic_df_fill)[20:30] = paste0("avgPSI_",colnames(junctionCountdf)[2:12])

for(i in 1:nrow(cryptic_df_fill)){
  ### LPSI 
  if(!is.na(cryptic_df_fill$left_junction[i])){
    # Identify inclusion Counts
    incJunc = unlist(as.list(str_split(cryptic_df_fill$left_junction[i],";")))
    incdf = junctionCountdf[which(junctionCountdf$Coordinates %in% incJunc),] %>% 
      dplyr::select(!Coordinates) %>% apply(2,function(x) sum(x,na.rm=T))
    incdf[is.na(incdf)] = 0
    
    # Identify exclusion counts
    excJunc = unlist(as.list(str_split(cryptic_df_fill$canonical_splice_junction[i],";")))
    excdf = junctionCountdf[which(junctionCountdf$Coordinates %in% excJunc),] %>% 
      dplyr::select(!Coordinates) %>% apply(2,function(x) sum(x,na.rm=T))
    excdf[is.na(excdf)] = 0
    
    # Calculate PSI
    psiCount_L = round((incdf/(excdf+incdf))*100,1)
    # cryptic_df_fill[i,20:30] = t(as.matrix(psiCount))
  }
  
  ### RPSI 
  if(!is.na(cryptic_df_fill$right_junction[i])){
    # Identify inclusion Counts
    incJunc = unlist(as.list(str_split(cryptic_df_fill$right_junction[i],";")))
    incdf = junctionCountdf[which(junctionCountdf$Coordinates %in% incJunc),] %>% 
      dplyr::select(!Coordinates) %>% apply(2,function(x) sum(x,na.rm=T))
    incdf[is.na(incdf)] = 0
    
    # Identify exclusion counts
    excJunc = unlist(as.list(str_split(cryptic_df_fill$canonical_splice_junction[i],";")))
    excdf = junctionCountdf[which(junctionCountdf$Coordinates %in% excJunc),] %>% 
      dplyr::select(!Coordinates) %>% apply(2,function(x) sum(x,na.rm=T))
    excdf[is.na(excdf)] = 0
    
    # Calculate PSI
    psiCount_R = round((incdf/(excdf+incdf))*100,1)
    # cryptic_df_fill[i,31:41] = t(as.matrix(psiCount))
  } 
  
  if(length(psiCount_L) > 0){
    psiCount_L <- as.numeric(str_replace_na(psiCount_L, replacement="0"))
    if(length(psiCount_R) > 0){
      psiCount_R <- as.numeric(str_replace_na(psiCount_R, replacement="0"))
      psiCount = (psiCount_L + psiCount_R)/2
      cryptic_df_fill[i,20:30] = t(as.matrix(psiCount))
    } else{
      cryptic_df_fill[i,20:30] = t(as.matrix(psiCount_L))
    } 
  } else{
    psiCount_R <- as.numeric(str_replace_na(psiCount_R, replacement="0"))
    cryptic_df_fill[i,20:30] = t(as.matrix(psiCount_R))
  }
  
}

cryptic_df_fill <- cryptic_df_fill %>% arrange(canonical_splice_junction) %>% arrange(cryptic_exon_coordinates) %>% arrange(gene_symbol) 
cryptic_df_fill$cryptic_exon_id <- paste0("CE", sprintf("%03d", 1:422))
# write.csv(cryptic_df_fill,"output_tables/251224_CrypticCoordinates.csv", row.names = F)

## Create new dataframe to populate with splice junction counts: SKIPTIC exons
skiptic_df_fill = skiptic_df
skiptic_df_fill[,6:16] = NA
colnames(skiptic_df_fill)[6:16] = paste0("avgPSI_",colnames(junctionCountdf)[2:12])

for(i in 1:nrow(skiptic_df_fill)){
  ### Identify inclusion Counts (skiptic)
  incJunc = skiptic_df_fill$skiptic_junction[i]
  incdf = junctionCountdf[which(junctionCountdf$Coordinates %in% incJunc),] %>% 
    dplyr::select(!Coordinates) %>% apply(2,function(x) sum(x,na.rm=T))
  incdf[is.na(incdf)] = 0
    
  ### Left canonical junction counts
  L_excJunc = skiptic_df_fill$left_canonical_junction[i]
  L_excdf = junctionCountdf[which(junctionCountdf$Coordinates %in% L_excJunc),] %>% 
    dplyr::select(!Coordinates) %>% apply(2,function(x) sum(x,na.rm=T))
  L_excdf[is.na(L_excdf)] = 0
  
  ### Right canonical junction counts
  R_excJunc = skiptic_df_fill$right_canonical_junction[i]
  R_excdf = junctionCountdf[which(junctionCountdf$Coordinates %in% R_excJunc),] %>% 
    dplyr::select(!Coordinates) %>% apply(2,function(x) sum(x,na.rm=T))
  R_excdf[is.na(R_excdf)] = 0
  
  # Calculate PSI
  psi_L = round((incdf/(L_excdf+incdf))*100,1)
  psi_R = round((incdf/(R_excdf+incdf))*100,1)
  avgPSI = (psi_L + psi_R)/2
  skiptic_df_fill[i,6:16] = t(as.matrix(avgPSI))
  
}

skiptic_df_fill$cryptic_exon_id <- paste0("CE", sprintf("%03d", 423:432))
# write.csv(skiptic_df_fill,"output_tables/251224_SkipticCoordinates.csv", row.names = F)

```


## Create Summary columns
Find rowwise means/avgs per condition, etc.

```{r SummaryColumns}
## Create one table with only gene symbol, ensembl id, ce id, and PSIs
cryptic_df_fill_subset <- cryptic_df_fill %>% 
  dplyr::select(c(gene_symbol,ensembl_id,cryptic_exon_id,starts_with("avgPSI_")))
skiptic_df_fill_subset <- skiptic_df_fill %>%
  dplyr::select(c(gene_symbol,ensembl_id,cryptic_exon_id,starts_with("avgPSI_")))

df <- rbind(cryptic_df_fill_subset,skiptic_df_fill_subset)
df[is.na(df)] = 0

# Add max
df$maxCtrl = apply(df %>% dplyr::select(contains("avgPSI_C")),1,function(x) max(x, na.rm=T))
df$maxTDP = apply(df %>% dplyr::select(contains("avgPSI_T")),1,function(x) max(x, na.rm=T))
df$maxDiff = df$maxTDP-df$maxCtrl

# write.csv(df,"output_tables/251223_cetable_psi.csv", row.names = F)
```

# Visualization

```{r pieChartFx}
### Pie chart of the proportions of types

pieChartFx = function(dfCol, naValue = F, cols = c("#262626","#D1D1D1")){
  ### Pie chart of the proportions of types
  propdf <- data.frame(table(dfCol,useNA="always"))
  colnames(propdf) = c("Value","Count")
  if(naValue){
    propdf$Value = c("Y","N")
  }
  
  # Compute percentages
  propdf$fraction = propdf$Count / sum( propdf$Count)
  
  # Compute a good label
  propdf$label <- propdf$Value

  # Label locations
  propdf$ymax = cumsum(propdf$fraction)
  propdf$ymin = c(0, head(propdf$ymax, n=-1))
  propdf$labelPosition <- (propdf$ymax + propdf$ymin) / 2

  # Plot proportion of each type
  p <- ggplot(propdf, aes(ymax=cumsum(propdf$fraction), ymin=c(0, head(cumsum(propdf$fraction), n=-1)), xmax=4, xmin=3, fill=label)) +
    geom_rect() +
    #geom_text( x=4.3, aes(y=labelPosition, label=label, color=label), size=3) + # x here controls label position (inner / outer)
    scale_color_manual(values = rep("black",nrow(propdf)))+
    scale_fill_manual(values = cols)+
    #scale_fill_brewer(palette=4) +
    coord_polar(theta="y",start = 90) +
    theme(text = element_text(family = "Helvetica"))+
    xlim(c(2, 4)) +
    theme_void() +
    theme(legend.position = "none")
  return(p)
}
```

```{r VisPieChart}
setwd("/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Scripts/NMDi_TDP43kd_repo/")
cryptic_df = read_delim("data/250601_CE_table.csv")

pieChartFx(cryptic_df$type, cols = c("#002D72","#69140E","#F2F3AE","#82735C"))
ggsave("output_fig/250602_CE_All.png",device = "png", 
       width = 30, height = 30, units = "cm",limitsize = FALSE)
table(cryptic_df$type)

pieChartFx(cryptic_df$`Inf. Ins.`, naValue = T)
ggsave("output_fig/250602_Pie_InFrame_CE.png",device = "png", 
       width = 30, height = 30, units = "cm",limitsize = FALSE)
table(cryptic_df$`Inf. Ins.`)

pieChartFx(cryptic_df$UTR, naValue = T)
ggsave("output_fig/250329_Pie_UTR_CE.png",device = "png", 
       width = 30, height = 30, units = "cm",limitsize = FALSE)
table(cryptic_df$UTR)

pieChartFx(cryptic_df$`Annotated Alternate Exon`, naValue = T)
ggsave("output_fig/250329_Pie_AnnotatedAlt_CE.png",device = "png", 
       width = 30, height = 30, units = "cm",limitsize = FALSE)
table(cryptic_df$`Annotated Alternate Exon`)

# NMD reveal
cryptic_df = read_delim("data/250601_CE_table.csv")
cryptic_df <- cryptic_df %>% mutate(NMDonly10pct = ifelse(avgT <= 10, "10pct", "TDPKDonly"),
                                    NMDonly10pct = ifelse(avgT <= 10, "10pct", "TDPKDonly"),
                                    NMDonly5pct = ifelse(avgT <= 5, "5pct", "TDPKDonly"),
                                    NMDonly1pct = ifelse(avgT <= 1, "1pct", "TDPKDonly")) %>%
  pivot_longer(cols=c("NMDonly10pct","NMDonly5pct","NMDonly1pct"), 
               names_to = "NMDonly", values_to = "category",
               names_prefix = "NMDonly") %>% 
  mutate(category = ifelse(is.na(category), NMDonly, category),
         category = ifelse(category == "TDPKDonly","TDPKDonly","NMDi"),
         NMDonly = factor(NMDonly, c("10pct","5pct","1pct"))) 
nmdOnlyCEs <- cryptic_df[which(cryptic_df$NMDonly == "1pct" & cryptic_df$category == "NMDi"),]

table(cryptic_df$NMDonly, cryptic_df$category)

cryptic_df %>% ggplot(aes(x=NMDonly,fill=category))+geom_bar()+
  theme_classic()+ xlab("") + ylab("# CEs") +
  theme(text = element_text(size = 14, family = "Helvetica"), 
        axis.ticks.x = element_blank(), 
        axis.title = element_text(size = 20, family = "Helvetica", color = "black")) +
  scale_y_continuous(expand = c(0, 0),limits = c(0,500))+
  scale_fill_manual(values = c("#D1D1D1","#262626"))
ggsave("output_fig/250602_NMDi_Effect_PctCutoffs.png",device = "png", 
       width = 30, height = 30, units = "cm",limitsize = FALSE)


### Plots to look at dPSI as a result of NMDi 
### Comparison to TDP-43 KD only
cryptic_df_dNMD <- cryptic_df[,c(1:19,42:54)] %>% 
  pivot_longer(20:30, names_to = "condition", names_prefix = "avg", values_to = "PSI")
cryptic_df_dNMD_avgT <- cryptic_df_dNMD %>% filter(condition == "T")
cryptic_df_dNMD_avgNMD <- cryptic_df_dNMD %>% filter(condition != "T") %>% 
  mutate(facetCol = condition)
cond <- unique(cryptic_df_dNMD$condition)

for(i in cond){
  cryptic_df_dNMD_avgNMD <- rbind(cryptic_df_dNMD_avgNMD, 
                                  cryptic_df_dNMD_avgT %>% mutate(facetCol = i))
  rm(i)
}

cryptic_df_dNMD_avgNMD %>% filter(facetCol %in% c("TXS","TX","TUS","TXU")) %>%
  mutate(facetCol = factor(facetCol, levels=c("TXS","TX","TUS","TXU"))) %>%
  ggplot(aes(x = condition, y = PSI)) +
  geom_line(aes(group=`ID #`), alpha = 0.07) +
  geom_boxplot(alpha=1, outliers = T, fill = c(rep(c("#444444","#D1D1D1"),4)), color = "black",
               outlier.color = "#444444",outlier.size = 5, linewidth = 3) +
  theme_classic()+ xlab("") + ylab("Average R/L PSI") +
  theme(text = element_text(size = 14, family = "Helvetica"), 
        axis.line.x = element_blank(),
        axis.ticks.x = element_blank(), 
        axis.title = element_text(size = 20, family = "Helvetica", color = "black"),
        plot.margin = margin(0.8, 0.8, 0.8, 0.8, "cm"),
        strip.background = element_rect(fill = "#02153D", colour = "#02153D"),
        strip.text = element_text(color="white"), panel.spacing = unit(0.3, "cm"))+
  #geom_hline(yintercept=0, linetype="dotted")+scale_y_continuous(breaks=c(seq(-50,100,25))) +
  facet_wrap(~ facetCol, scales = "free_x", ncol = 4) 
ggsave("output_fig/251212_NMDi_Effect_Paired.png",device = "png", 
       width = 50, height = 40, units = "cm",limitsize = FALSE)


cryptic_df_dNMD_avgNMD %>% filter(facetCol %in% c("TXS","TX","TUS","TXU")) %>%
  mutate(condition = factor(condition, levels=c("T","TXS","TX","TUS","TXU"))) %>%
  ggplot(aes(x = condition, y = PSI)) +
  geom_line(aes(group=`ID #`, color=`ID #`), alpha = 0.3,linewidth=0.5) +
  geom_boxplot(alpha=1, outliers = T, fill = "#444444", color = "black",
               outlier.color = "#444444",outlier.size = 1) +
  scale_color_viridis_d(option = "viridis")+
  theme_classic()+ xlab("") + ylab("Average R/L PSI") +
  theme(text = element_text(size = 25, family = "Helvetica"), legend.position = "none",
        axis.line.x = element_blank(),
        axis.ticks.x = element_blank(), 
        axis.title = element_text(size = 20, family = "Helvetica", color = "black"),
        plot.margin = margin(0.8, 0.8, 0.8, 0.8, "cm"),
        strip.background = element_rect(fill = "#02153D", colour = "#02153D"),
        strip.text = element_text(color="white"), panel.spacing = unit(0.3, "cm"))+
  geom_hline(yintercept=0, linetype="dotted")+scale_y_continuous(breaks=c(seq(-50,100,25))) +
    stat_compare_means(label = "p.signif", method = "t.test", ref.group = "T")
ggsave("output_fig/250506_NMDi_Effect_Stats.png",device = "png", 
       width = 30, height = 30, units = "cm",limitsize = FALSE)

# 1% NMD CEs increase
cryptic_df_dNMD_avgNMD %>% filter(facetCol %in% c("TXS","TX","TUS","TXU")) %>%
  mutate(condition = factor(condition, levels=c("T","TXS","TX","TUS","TXU"))) %>%
  filter(`ID #` %in% nmdOnlyCEs$`ID #`) %>%
  ggplot(aes(x = condition, y = PSI)) +
  #geom_line(aes(group=`ID #`), alpha = 0.3,linewidth=0.5) +
  geom_boxplot(alpha=1, outliers = T, fill = "#444444", color = "black",
               outlier.color = "#444444",outlier.size = 1) +
  scale_color_viridis_d(option = "viridis")+
  theme_classic()+ xlab("") + ylab("Average R/L PSI") +
  theme(text = element_text(size = 25, family = "Helvetica"), legend.position = "none",
        axis.line.x = element_blank(),
        axis.ticks.x = element_blank(), 
        axis.title = element_text(size = 20, family = "Helvetica", color = "black"),
        plot.margin = margin(0.8, 0.8, 0.8, 0.8, "cm"),
        strip.background = element_rect(fill = "#02153D", colour = "#02153D"),
        strip.text = element_text(color="white"), panel.spacing = unit(0.3, "cm"))+
  geom_hline(yintercept=0, linetype="dotted")+scale_y_continuous(breaks=c(seq(-50,100,25))) +
    stat_compare_means(label = "p.signif", method = "t.test", ref.group = "T")

### chr/mbp graph 
chrLength <- read_delim("data/GRch38sequence_report.tsv") %>% 
  filter(`Assembly-unit accession`=="Primary Assembly", 
         `Sequence name` %in% c(seq(1,22,1),"X","Y"))
cryptic_df = read_delim("output_tables/250329_CrypticCoordinates_Vis.csv")
cryptic_df <- cryptic_df %>% mutate(diffCUS = avgCUS-avgCtrl,
                      diffCX = avgCX-avgCtrl,
                      diffCXS = avgCXS-avgCtrl,
                      diffCXU = avgCXU-avgCtrl,
                      diffT = avgT-avgT,
                      diffTUS = avgTUS-avgT,
                      diffTX = avgTX-avgT,
                      diffTXS = avgTXS-avgT,
                      diffTXU = avgTXU-avgT)  %>% 
  mutate(chr = gsub("chr","",
                    str_split_i(cryptic_df$`Canonical Splice Junction`,":",1))) %>%
  mutate(chr = factor(chr, levels=c(seq(1,22,1),"X","Y")))
 # write.csv(cryptic_df,"output_tables/250329_CrypticCoordinates_Vis.csv", row.names = F) # Overwrites the other file

histcryptic <- data.frame(table(cryptic_df$chr))
histcryptic <- left_join(histcryptic, chrLength, 
                         by = join_by("Var1" == "Chromosome name")) %>%
  mutate(chrM = Freq/`Seq length`, Var1 = factor(Var1, levels = c(seq(1,22,1),"X","Y"))) 

histcryptic %>% 
  ggplot(aes(x=Var1, y= chrM*(10^6)))+geom_col() +theme_classic()+ xlab("") +
  ylab("CEs/Mbp") +scale_y_continuous(expand = c(0, 0)) +
  theme(text = element_text(size = 25, family = "Helvetica", color = "black"),
        axis.line.x = element_blank(),
        axis.ticks.x = element_blank(), 
        axis.title = element_text(size = 20, family = "Helvetica", color = "black"),
        plot.margin = margin(0.8, 0.8, 0.8, 0.8, "cm"),
        strip.background = element_rect(fill = "#02153D", colour = "#02153D"),
        strip.text = element_text(color="white"), panel.spacing = unit(0.3, "cm"))
ggsave("output_fig/250506_chrPmbp.png",device = "png", 
       width = 50, height = 25, units = "cm",limitsize = FALSE)
```

```{r HighExp}
setwd("/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Scripts/NMDi_TDP43kd_repo/")
cryptic_df = read_delim("data/250601_CE_table.csv")
nauc <- read_delim("/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Projects/NMD/NotGitHub/data/gtex_nauc.csv")

## High expressors
cryptic_df_high <- cryptic_df %>% filter(maxTDP > 70, maxCtrl < 10) 
# write.csv(cryptic_df_high,"output_tables/250602_CrypticCoordinates_HighExp.csv")

pieChartFx(cryptic_df_high$type, cols = c("#002D72","#69140E","#F2F3AE","#82735C"))
table(cryptic_df_high$type)


### Plots to look at dPSI as a result of NMDi 
### Comparison to TDP-43 KD only
cryptic_df_dNMD <- cryptic_df_high[,c(1:18,42:54)] %>% 
  pivot_longer(20:30, names_to = "condition", names_prefix = "avg", values_to = "PSI")
cryptic_df_dNMD_avgT <- cryptic_df_dNMD %>% filter(condition == "T")
cryptic_df_dNMD_avgNMD <- cryptic_df_dNMD %>% filter(condition != "T") %>% 
  mutate(facetCol = condition)
cond <- unique(cryptic_df_dNMD$condition)

for(i in cond){
  cryptic_df_dNMD_avgNMD <- rbind(cryptic_df_dNMD_avgNMD, 
                                  cryptic_df_dNMD_avgT %>% mutate(facetCol = i))
  rm(i)
}

cryptic_df_dNMD_avgNMD %>% filter(facetCol %in% c("TXS","TX","TUS","TXU")) %>%
  mutate(condition = factor(condition, levels=c("T","TXS","TX","TUS","TXU"))) %>%
  ggplot(aes(x = condition, y = PSI)) +
  #geom_line(aes(group=`ID #`), alpha = 0.3,linewidth=0.5) +
  geom_boxplot(alpha=1, outliers = T, fill = "#444444", color = "black",
               outlier.color = "#444444",outlier.size = 1) +
  scale_color_viridis_d(option = "viridis")+
  theme_classic()+ xlab("") + ylab("Average R/L PSI") +
  theme(text = element_text(size = 25, family = "Helvetica"), legend.position = "none",
        axis.line.x = element_blank(),
        axis.ticks.x = element_blank(), 
        axis.title = element_text(size = 20, family = "Helvetica", color = "black"),
        plot.margin = margin(0.8, 0.8, 0.8, 0.8, "cm"),
        strip.background = element_rect(fill = "#02153D", colour = "#02153D"),
        strip.text = element_text(color="white"), panel.spacing = unit(0.3, "cm"))+
  geom_hline(yintercept=0, linetype="dotted")+scale_y_continuous(breaks=c(seq(-50,100,25))) +
    stat_compare_means(label = "p.signif", method = "wilcox.test", ref.group = "T")

## HEAT MAPS

# High expressor gene expression across tissues
cryptic_df_high <- cryptic_df %>% filter(maxTDP > 70, maxCtrl < 10) 

nauc_highCE <- nauc %>% filter(gene_id %in% cryptic_df_high$`Ensembl ID`) %>% 
  pivot_longer(3:58, names_to = "tissue", values_to = "nauc") %>%
  mutate(nauc = ifelse(nauc >=10, 10, nauc))

nauc_highCE_mat <- nauc_highCE[-1] %>% pivot_wider(names_from="tissue", values_from="nauc") %>% column_to_rownames(var="gene_symbol")

pheatmap(t(as.matrix(nauc_highCE_mat)), 
         color = viridis(length(seq(0,10,0.1)),option="D"),
         border_color = NA, 
         cluster_rows = F,
         show_rownames = F,
         treeheight_row = 0, 
         treeheight_col = 0,
         breaks = seq(0,10,0.1)
         )

# Create matching PSI graph
 

orderGenes <- pheatmap(t(as.matrix(nauc_highCE_mat)), 
         color = viridis(length(seq(0,10,0.1)),option="D"),
         border_color = NA, 
         cluster_rows = F,
         show_rownames = F,
         treeheight_row = 0, 
         treeheight_col = 0,
         breaks = seq(0,10,0.1)
         )
crypticHighOrder <- data.frame(gene = orderGenes$tree_col$labels[orderGenes$tree_col$order])

# Choose highest PSI CE for a gene with multiple in high expressors
maxPSI <- cryptic_df_high %>% filter(`Ensembl ID` %in% nauc_highCE$gene_id) %>%
  group_by(`Gene Symbol`) %>%
  summarize(max = max(maxTDP)) 

maxCE <- c()
for(i in 1:nrow(maxPSI)){
  maxCE <- append(maxCE, which(cryptic_df_high$`Gene Symbol`== maxPSI$`Gene Symbol`[i] & cryptic_df_high$maxTDP== maxPSI$max[i]))
}

maxPSI <- cryptic_df_high[maxCE,c(1:4,42:54)]

# Still two duplicates due to same max - rmove the ones at a lower PSI after TDP kd only
maxPSI <- maxPSI %>% filter(!(`ID #` %in% c("CE140","CE360"))) 
maxPSI <- maxPSI[,c(1,5,8,7,6,9,10,13,12,11,14)]
colnames(maxPSI)[1] = "gene"

# MERGE WITH ORDERED DF
maxPSI$gene[15] = "CTB-50L17.10"
maxPSI$gene[12] = "C20orf194"

crypticHighOrder <- left_join(crypticHighOrder, maxPSI, by = "gene") %>% 
  column_to_rownames("gene")
crypticHighOrder[is.na(crypticHighOrder)] = 0

pheatmap(t(as.matrix(crypticHighOrder)), 
         color = viridis(length(seq(0,100,1)),option="F"),
         border_color = NA, 
         cluster_rows = F,
         cluster_cols = F,
         show_rownames = F,
         treeheight_row = 0, 
         treeheight_col = 0,
         breaks = seq(0,100,1)
         )


```

## NMDi-revealed CEs expression and PSI
```{r NMDiCE}
setwd("/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Scripts/NMDi_TDP43kd_repo/")
cryptic_df = read_delim("data/250601_CE_table.csv")
nauc <- read_delim("/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Projects/NMD/NotGitHub/data/gtex_nauc.csv")

## Subset to CEs with PSI <= 1% after TDP-43 KD only
# all ctrl conditions < 5%
cryptic_df <- cryptic_df %>% mutate(NMDonly = ifelse(avgT <= 1, ifelse(maxCtrl < 10, "Y", "N"), "N"))
nmdOnlyCEs <- cryptic_df[which(cryptic_df$NMDonly == "Y"),]

## Table Information
nmdOnlyCEs_df <- cryptic_df %>% 
  mutate(NMDonly = ifelse(avgT <= 10, ifelse(maxCtrl < 10, "10pct", "N"), "N"),
         NMDonly = ifelse(avgT <= 5, ifelse(maxCtrl < 10, "5pct", NMDonly), NMDonly),
         NMDonly = ifelse(avgT <= 1, ifelse(maxCtrl < 10, "1pct", NMDonly), NMDonly)) %>%
  filter(NMDonly != "N")

# write.csv(nmdOnlyCEs_df[,c(1:7,9,42:55)],"output_tables/251215_NMD_CrypticCoordinates.csv", row.names = F)

nauc_NMD <- nauc %>% filter(gene_id %in% nmdOnlyCEs$`Ensembl ID`) %>%
  dplyr::select(-gene_id) %>% 
  pivot_longer(2:57,names_to = "tissue", values_to = "nauc") %>%
  mutate(nauc = ifelse(nauc > 10, 10, nauc)) %>%
  pivot_wider(names_from = "tissue", values_from = "nauc") %>% 
  column_to_rownames("gene_symbol") 

# Plot nauc of all NMD-sensitive CEs
#NAUC
pheatmap(t(as.matrix(nauc_NMD)), 
         cluster_cols = T,
         cluster_rows = F,
         fontsize = 12,
         color = viridis(100,option="D"), 
         legend_breaks = c(0,5,10),
         border_color = NA,
         display_numbers = F,
         treeheight_col = 0)

# PSI
cryptic_df_plot <- cryptic_df %>% 
  filter(`Gene Symbol` %in% c("CYFIP2", "NRXN2", "GAK", "NADSYN1", "ENAH", "PBX1")) %>%
  dplyr::select(c(`Gene Symbol`, `ID #`,starts_with("avg"))) %>%
  filter(!(`ID #` %in% c("CE092", "CE093","CE267"))) %>% # MANUAL SELECTION OF STRONGEST CE 
  dplyr::select(-`ID #`) %>%
  pivot_longer(2:11, names_to = "cond", values_to = "psi") %>%
  pivot_wider(names_from = cond, values_from = psi, names_sort = T) %>%
  column_to_rownames("Gene Symbol")

#Re-order
cryptic_df_plot <- cryptic_df_plot[c(1,5,3,4,2,6),c(1, 4, 3, 2, 5, 6, 9, 8, 7, 10)]

# Sets the minimum (0), the maximum (15), and the increasing steps (+1) for the color scale
# Note: if some of your genes are outside of this range, they will appear white on the heatmap
breaksList = seq(0, 30, by = 0.1)

pheatmap(t(as.matrix(cryptic_df_plot)), 
         cluster_cols = F,
         cluster_rows = F,
         fontsize = 12,
         color = viridis(length(breaksList),option="F"), 
         legend_breaks = c(0,10,20,30),
         border_color = NA,
         display_numbers = F,
         treeheight_col = 0,
         breaks= breaksList)

```

