---
title: "250410_GEX"
author: "Irika Sinha"
date: "2025-04-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Purpose

The purpose of this script is to take the file with manually identified cryptic exon coordinates and use those coordinates to identify left and right junction PSIs in the different conditions for i3N. 

## Packages

```{r packages, warning=FALSE, message=FALSE,}
#update.packages("BiocManager")
library("BiocManager")
packages <- c("tidyverse", "ggplot2", "ggpubr", "ggstats","BiocManager",
              "viridis","gghighlight", "ggrepel","apeglm", "DESeq2", "knitr",
              "biomaRt", "readxl")

for(i in packages){
  if(!(i %in% rownames(installed.packages()))){
    BiocManager::install(pkgs = i, ask = F)
  } else{
    #BiocManager::install(pkgs = i, ask = F, update=T)
  }
  library(package = i, character.only = T, quietly = T)
  rm(i)
}

rm(packages)

## List Package Versions
versions = installed.packages()[names(sessionInfo()$otherPkgs), "Version"]
data.frame(versions)
rm(versions)

```

## DEG Analysis
Nextflow RNA-seq pipeline used to generate DESeq2 objects.

```{r addMeta, eval=F}
setwd("/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Scripts/NMDi_TDP43kd_repo/")
# Load data
load("/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Projects/NMD/NotGitHub/data/nmdi_all_deseq2.dds.RData")

# Add metadata
colnames(dds@colData)[which(colnames(dds@colData) == "Group1")] = "Treatment"
colnames(dds@colData)[which(colnames(dds@colData) == "Group2")] = "Rep"
dds@colData$NMDi = as.factor(c(rep("WT",2),rep("KD",8),rep("WT",4),rep("KD",8)))
dds@colData$NMDi = factor(dds@colData$NMDi, levels =c("WT","KD"))
dds@colData$tdp = as.factor(c(rep("WT",12),rep("KD",10)))
dds@colData$tdp = factor(dds@colData$tdp, levels =c("WT","KD"))
dds@colData$seqDate = as.factor(c(rep("batch1",10),rep("batch2",12)))
design(dds) <- ~ tdp + NMDi + seqDate

# saveRDS(dds, "data/nmdi_all_annotated_deseq2.dds.RDS")

```

## Run DESeq2 
Used DESeq2 vignette to write analysis script: https://www.bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html

First run on the orginal dds to identify gene expression changes without adjusting for PSI.
```{r DESeq2_org, eval = F}
dds <- readRDS("data/nmdi_all_annotated_deseq2.dds.RDS")

## Subset to samples of interest
## Pre-filtering - there are 4 samples
dds <- dds[,11:14]
dds <- dds[rowSums(counts(dds)) >= 10,] # at least 10 total reads across all samples
dds_counts <- data.frame(dds@assays@data@listData$counts) # Extract counts for later

# saveRDS(dds_counts, "data/dds_counts_org.RDS")

## DESeq2 Analysis
# Add experiment design
# LFC shrinkage for vis + Extract results
dds$tdp <- factor(dds$tdp, levels = c("WT","KD"))
design(dds) <- ~ tdp
dds <- DESeq(dds)
res <- lfcShrink(dds, coef="tdp_KD_vs_WT", type="apeglm")

# saveRDS(res, "data/dds_res_org.RDS")

```

Next, use PSI values to raw counts for CE-related downregulation.
functional transcript = counts * (1-PSI)

```{r DESeq2_adj, eval = F}
cryptic_df = read_delim("data/250601_CE_table.csv")
adj_count <- readRDS("data/dds_counts_org.RDS")

## Create new CE df which accounts for multiple CEs for a single gene
# Choose higher avgPSI for a gene with multiple CEs
cryptic_df_PSImax <- cryptic_df[,-c(7:17)] %>% 
  group_by(`Gene Symbol`, `Ensembl ID`) %>% 
  summarize(avgT = max(avgPSI_sgTARDBP)) %>% # For gene with multiple CEs, choose higher PSI
  mutate(avgT = ifelse(!is.na(avgT),avgT, 0)) %>% # If there are no counts for this gene, the PSI is 0
  left_join(cryptic_df %>% dplyr::select(`Gene Symbol`,type,avgPSI_sgTARDBP,maxPSI_sgTARDBP), 
            by = join_by(`Gene Symbol`==`Gene Symbol`,avgT==avgPSI_sgTARDBP)) %>%
  ungroup() %>%
  mutate(pct_transcript = 1-avgT/100, type = tolower(type)) %>%
  distinct()

# For each gene with a cryptic exon, adjust the TDPKD value to reflect effective down-regulation of the gene due to CE inclusion
for(i in 1:nrow(cryptic_df_PSImax)){
  n = which(rownames(adj_count) == cryptic_df_PSImax$`Ensembl ID`[i]) # Identify row in counts for CE
  pct_transcript = cryptic_df_PSImax$pct_transcript[i] 
  
  # Adjust counts
  adj_count$TDPKD_1[n] = as.integer(round((adj_count$TDPKD_1[n])*pct_transcript,0))
  adj_count$TDPKD_2[n] = as.integer(round((adj_count$TDPKD_2[n])*pct_transcript,0))
  
  rm(i, n,pct_transcript)
}

## Create DESeq2 object and do analysis as before
# Extract results
dds_adj <- DESeqDataSetFromMatrix(countData = as.matrix(adj_count),
                              colData = dds@colData,
                              design = ~ tdp)

dds_adj <- DESeq(dds_adj)
res_adj <- lfcShrink(dds_adj, coef="tdp_KD_vs_WT", type="apeglm")

# saveRDS(res_adj, "data/dds_res_adj.RDS")

```

## Create results dataframes for plotting

```{r DESeq2_df, eval=F}

res_org <- readRDS("data/dds_res_org.RDS")
res_adj <- readRDS("data/dds_res_adj.RDS")

## Create a df with all results
res_df <- data.frame(res_org) %>% rownames_to_column("ensg") %>%
  mutate(logp_org = -log10(padj)) %>%
  dplyr::select(ensg,log2FoldChange,logp_org) %>% 
  dplyr::rename(lfc_org = log2FoldChange)
res_adj_df <- data.frame(res_adj) %>% rownames_to_column("ensg") %>% 
  mutate(logp_adj = -log10(padj)) %>%
  dplyr::select(ensg,log2FoldChange,logp_adj) %>% 
  dplyr::rename(lfc_adj = log2FoldChange)

res_df <- full_join(res_df, res_adj_df, by = "ensg")

## Add gene names
mart <- useEnsembl(biomart = "ensembl", dataset = "hsapiens_gene_ensembl")
info <- getBM(attributes = c("ensembl_gene_id", "external_gene_name","strand", "gene_biotype"), 
              filters = "ensembl_gene_id", values = res_df$ensg, mart = mart) %>%
  filter(gene_biotype %in% c("protein_coding")) %>%
  dplyr::select(-c(gene_biotype))

## Filter genes to protein-coding genes
res_df <- left_join(info, res_df, by = join_by("ensembl_gene_id" == "ensg"))

# Add empty cols for CE type
res_df <- res_df %>% mutate(ce_type = NA, tdpkd_psi = NA, tdpkd_nmdi_psi = NA)

for(i in 1:nrow(cryptic_df_PSImax)){
  n = which(res_df$ensembl_gene_id == cryptic_df_PSImax$`Ensembl ID`[i])
  res_df$ce_type[n] = cryptic_df_PSImax$type[i]
  res_df$tdpkd_psi[n] = cryptic_df_PSImax$avgT[i]
  res_df$tdpkd_nmdi_psi[n] = cryptic_df_PSImax$maxPSI_sgTARDBP[i]
  
  rm(i,n)
}

# saveRDS(res_df, "output_tables/251216_NMDi_bulkseq_res.RDS")

```

## Create Plots!! Yay

This is the unadjusted volcano plot for gene expression
```{r plot_org, fig.align = "center", out.width="100%", fig.dpi=300, fig.asp=0.6}
setwd("/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Scripts/NMDi_TDP43kd_repo")

# Plot log2FC vs -log10Padj with CEs with abs(lfc) > 1 and p.adj < 10^-20 named on plot
res_df %>% mutate(logp_org = ifelse(is.infinite(logp_org) | logp_org > 50, 50, logp_org), # Set max limit for -log10P
                  lfc_org = ifelse(lfc_org < -5,-5, ifelse(lfc_org > 5, 5, lfc_org)), # Set limits for log2FC
                  gene_label = ifelse(logp_org>20 & abs(lfc_org) > 1,external_gene_name,"")) %>% # ID labeled CEs
  ggplot(aes(x=lfc_org,y=logp_org, color=factor(ce_type)))+ 
  geom_point(alpha=0.8,size=3) + theme_classic() + 
  gghighlight(!is.na(ce_type),unhighlighted_params = list(alpha=0.5, size=2))+ # Color points by CE type
  geom_label_repel(aes(label = gene_label), 
                   box.padding = 0.5, max.overlaps = Inf, fontface = "bold",
                   fill = alpha(c("white"),0.5), label.size = NA)+
  scale_color_manual(values=c("#002D72","#69140E","#ffc000","#82735C"))+
  theme(text = element_text(family="Helvetica"))+
  ylim(0,60) + ylab("-log10P") + xlab("log2FC")

```

Adjust the gene expression for CE-induced downregulation.

```{r plot_adj, fig.align = "center", out.width="100%", fig.dpi=300, fig.asp=0.6}

# Plot log2FC vs -log10Padj with CEs with abs(lfc) > 1 and p.adj < 10^-20 named on plot
res_df %>% mutate(logp_adj = ifelse(is.infinite(logp_adj) | logp_adj > 50, 50, logp_adj), # Set max limit for -log10P
                  lfc_adj = ifelse(lfc_adj < -5,-5, ifelse(lfc_adj > 5, 5, lfc_adj)), # Set limits for log2FC
                  gene_label = ifelse(logp_adj>20 & abs(lfc_adj) > 1,external_gene_name,"")) %>% # ID labeled CEs
  ggplot(aes(x=lfc_adj,y=logp_adj, color=factor(ce_type)))+ 
  geom_point(alpha=0.8,size=3) + theme_classic() + 
  gghighlight(!is.na(ce_type),unhighlighted_params = list(alpha=0.5, size=2))+ # Color points by CE type
  geom_label_repel(aes(label = gene_label), 
                   box.padding = 0.5, max.overlaps = Inf, fontface = "bold",
                   fill = alpha(c("white"),0.5), label.size = NA)+
  scale_color_manual(values=c("#002D72","#69140E","#ffc000","#82735C"))+
  theme(text = element_text(family="Helvetica"))+
  ylim(0,60) + ylab("-log10P") + xlab("log2FC")
```

Plot impact of PSI adjustment

```{r plot_log2FCdiff, fig.align = "center", out.width="100%", fig.dpi=300, fig.asp=0.6}

# Graph CE-adjusted log2FC and unadjusted log2FC
res_df %>% filter(!is.na(ce_type)) %>% # Only plot genes with CEs
  mutate(lfc_org = ifelse(lfc_org < -2,-2, ifelse(lfc_org > 2, 2, lfc_org)), # Set limits for log2FC
         lfc_adj = ifelse(lfc_adj < -2,-2, ifelse(lfc_adj > 2, 2, lfc_adj)), # Set limits for log2FC
         gene_label = ifelse(tdpkd_psi > 50,external_gene_name,"")) %>% # Label points for genes with PSI >= 50% 
  ggplot(aes(x=lfc_org,y=lfc_adj))+ 
  geom_point(alpha=0.8,size=5, color = "#69140E") + theme_classic() + 
  geom_abline(intercept = 0, slope = 1, linetype = "dotted",linewidth = 1)+
  geom_label_repel(aes(label = gene_label), 
                   box.padding = 0.5, max.overlaps = Inf,fontface = "bold",
                   fill = alpha(c("white"),0.8), label.size = NA)+
  theme(text = element_text(family="Helvetica"))+ylim(-2.2,2)+
  labs(x = "log2FC (original)", y = "log2FC (psi-adj)", 
       title = "CE inclusion leads to increased downregulation")

```

## NMD sensitivity of CE vs l2fc

```{r plot_NMD, fig.align = "center", out.width="75%", fig.dpi=300, fig.asp=1}
## Look at NMD sensitive CEs

res_df %>% filter(!is.na(ce_type)) %>% # Only plot genes with CEs
  mutate(lfc_adj = ifelse(lfc_adj < -5,-5, ifelse(lfc_adj > 5, 5, lfc_adj)), # Set limits for log2FC
         tdpkd_psi = ifelse(is.na(tdpkd_psi), 0, tdpkd_psi),
         tdpkd_nmdi_psi = ifelse(is.na(tdpkd_nmdi_psi), 0, tdpkd_nmdi_psi),
         l2psi = log2((tdpkd_nmdi_psi+0.01)/(tdpkd_psi+0.01)),
         gene_label = ifelse(tdpkd_psi > 50,external_gene_name,""),
         gene_NMD = ifelse(tdpkd_psi < 10, ifelse(tdpkd_psi < 5, "5pct", "10pct"), "TDP-43 KD only")) %>% # Label points for genes with PSI >= 50% 
  ggplot(aes(x=lfc_adj,y=l2psi))+ 
  geom_point(alpha=0.8,size=5, aes(fill = factor(gene_NMD)), shape = 21, color = "#363636") + theme_classic() + 
  gghighlight(gene_NMD != "TDP-43 KD only",unhighlighted_params = list(alpha=0.8, size=5, shape = 16))+
  scale_fill_manual(values=c("#69140E","#e5a4cb"))+
  theme(text = element_text(family="Helvetica"))+
  geom_vline(xintercept = 0, linetype = "dotted", linewidth = 1)+
  xlim(-5,2.5)+ylim(0,6)+
  labs(x = "log2FC (psi-adj)", y = "log2PSI (nmdi/tdponly)", 
       title = "Increased nmdi of ce moderately correlated with decreased l2fc")

# Graph CE-adjusted log2FC and unadjusted log2FC
res_df %>% filter(!is.na(ce_type)) %>% # Only plot genes with CEs
  mutate(lfc_org = ifelse(lfc_org < -2,-2, ifelse(lfc_org > 2, 2, lfc_org)), # Set limits for log2FC
         lfc_adj = ifelse(lfc_adj < -2,-2, ifelse(lfc_adj > 2, 2, lfc_adj)), # Set limits for log2FC
         tdpkd_psi = ifelse(is.na(tdpkd_psi), 0, tdpkd_psi),
         tdpkd_nmdi_psi = ifelse(is.na(tdpkd_nmdi_psi), 0, tdpkd_nmdi_psi),
         gene_NMD = ifelse(tdpkd_psi < 10, ifelse(tdpkd_psi < 5, "5pct", "10pct"), "TDP-43 KD only"),
         gene_label = ifelse(tdpkd_psi > 50,external_gene_name,"")) %>% # Label points for genes with PSI >= 50% 
  ggplot(aes(x=lfc_adj,y=tdpkd_nmdi_psi))+ 
  geom_point(alpha=0.8,size=5, aes(fill = factor(gene_NMD)),  shape = 21, color = "#363636") +   theme_classic() + 
  gghighlight(gene_NMD != "TDP-43 KD only",unhighlighted_params = list(alpha=0.8, size=5, shape = 16))+
  scale_fill_manual(values=c("#69140E","#e5a4cb"))+
  geom_vline(xintercept = 0, linetype = "dotted", linewidth = 1)+
  geom_label_repel(aes(label = gene_label), 
                   box.padding = 0.5, max.overlaps = Inf,fontface = "bold",
                   fill = alpha(c("white"),0.8), label.size = NA)+
  theme(text = element_text(family="Helvetica"))

```

## qPCR targets
```{r qpcr, fig.align = "center", out.width="75%", fig.dpi=300, fig.asp=0.6}
## Load files
qpcr_df <- readxl::read_xlsx("data/250905_qPCR_RNAs.xlsx")
qpcr_genes <- unique(qpcr_df$gene) 
qpcr_genes <- qpcr_genes[which(!(qpcr_genes %in% c("Ctrl","TARDBP")))]

# Plot log2FC vs -log10Padj
# Highlight qPCR genes
res_df %>% mutate(logp_adj = ifelse(is.infinite(logp_adj) | logp_adj > 50, 50, logp_adj), # Set max limit for -log10P
                  lfc_adj = ifelse(lfc_adj < -5,-5, ifelse(lfc_adj > 5, 5, lfc_adj)), # Set limits for log2FC
                  gene_label = ifelse(external_gene_name %in% qpcr_genes,external_gene_name,"")) %>% # ID labeled CEs
  filter(!is.na(logp_adj)) %>% ungroup() %>% 
  ggplot(aes(x=lfc_adj,y=logp_adj, color=factor(ce_type)))+ 
  geom_point(alpha=0.8,size=3) + theme_classic() + 
  gghighlight(gene_label != "",unhighlighted_params = list(alpha=0.5, size=2))+ # Color points by CE type
  geom_label_repel(aes(label = gene_label), 
                   box.padding = 0.5, max.overlaps = Inf, fontface = "bold",
                   fill = alpha(c("white"),0.5), label.size = NA)+
  scale_color_manual(values=c("#002D72","#69140E","#ffc000","#82735C"))+
  theme(text = element_text(family="Helvetica"))+
  ylim(0,60) + ylab("-log10P") + xlab("log2FC")


```
