---
title: "251124_SnapMine_MES"
author: "Irika Sinha"
date: "2025-11-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Purpose

The purpose of this script is use SnapMine to check PSIs of CEs in public RNA-Seq datasets

## Packages

SnapMine package is still in development but is functional. It can be downloaded from GitHub.

```{r packages, message = FALSE, warning = FALSE}
#update.packages("BiocManager")
library("BiocManager")
packages <- c("tidyverse","ggplot2","data.table","biomaRt", "pheatmap","viridis", "progress", "readxl","makeunique")

for(i in packages){
  if(!(i %in% rownames(installed.packages()))){
    BiocManager::install(pkgs = i, ask = F)
  } else{
    BiocManager::install(pkgs = i, ask = F, update=T)
  }
  library(package = i, character.only = T, quietly = T)
  rm(i)
}

rm(packages)

## Install SnapMine package from GitHub
remotes::install_github("irikas/SnapMine_Paper", subdir = "snapmine")
library(snapmine)

## List Package Versions
versions = installed.packages()[names(sessionInfo()$otherPkgs), "Version"]
data.frame(versions)
rm(versions)

```

## Load files and run Snaptron query
Copied from introduction vignette!

```{r files, eval=F}
## Load df
df <- fread("/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Scripts/NMDi_TDP43kd_repo/data/251126_CEdf.csv")

## Original df did not have strand info so need to use biomaRt
## Update df
mart <- useEnsembl(biomart = "ensembl", dataset = "hsapiens_gene_ensembl")
info <- getBM(attributes = c("ensembl_gene_id", "strand"), values = df$`Ensembl ID`, mart = mart)
df <- df %>%
  left_join(info, by = join_by(`Ensembl ID` == ensembl_gene_id)) %>%
  mutate(strand = ifelse(strand == 1, "+", "-"))
rm(mart, info)

## Create info_df file
info_df <- data.frame(
  novel_junc_id = paste(df$`Gene Symbol`, df$`ID #`, sep = "_"),
  compilation = "sra_human",
  strand = df$strand,
  novel_junc_left_coord = df$LeftJunction,
  novel_junc_right_coord = df$RightJunction,
  canon_junc_coord = df$`Canonical Splice Junction`
)
```

```{r query, eval = F}
## Run query
# Run twice for long and wide dfs.
# Could also run without flattening - this would lead to a list of dfs and then both flatten functions could be done manually using [flatten_counts]
result_df_long <- calcPSI_bulk(info_df, flatten_long = T)
result_df_wide <- calcPSI_bulk(info_df, flatten_wide = T)
# saveRDS(result_df_long, "/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Scripts/250325_NMDi/output_tables/251204_snapmine_results.RDS")
# saveRDS(result_df_wide, "/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Scripts/250325_NMDi/output_tables/251204_snapmine_result_df_wide.RDS")
```

```{r load_df, include = F}
df <- fread("/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Scripts/NMDi_TDP43kd_repo/data/251126_CEdf.csv")
info_df <- readRDS("/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Scripts/NMDi_TDP43kd_repo/data/251206_SnapMine_info_df.RDS")
# result_df_long <- readRDS("/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Projects/NMD/NotGitHub/output_tables/251204_snapmine_results.RDS")
result_df_wide <- readRDS("/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Projects/NMD/NotGitHub/output_tables/251204_snapmine_result_df_wide.RDS")
result_df_wide[is.na(result_df_wide)] = 0
```

## Visualize CE inclusion

```{r Visual, fig.align = "c", out.width="100%", fig.dpi=300, fig.asp=1}
## Subset to samples with avgPSI > 15 for at least one CE
result_df_wide_subset <- result_df_wide %>% 
  mutate(maxPSI = apply(result_df_wide[, -1],
                        MARGIN = 1, function(x) max(x, na.rm = T))) %>%
  filter(maxPSI > 5) %>% dplyr::select(-maxPSI) 

## Subset to avgPSI > 5 for at least 10% tested CEs
result_df_wide_subset <- result_df_wide_subset%>%
  mutate(testFilt = apply(result_df_wide_subset[, -1],
                     MARGIN = 1,
                     function(x) length(which(x > 5)))) %>%
  filter(testFilt > round(0.10 * nrow(info_df), 0)) %>%
  dplyr::select(-testFilt) %>%
  column_to_rownames(var = "sampleID")

## Remove CEs with no values higher than 10%
colKeep <- apply(X = result_df_wide_subset, MARGIN = 2, FUN = function(x) max(x) > 10)
result_df_wide_subset <- result_df_wide_subset[,colKeep]

pheatmap(as.matrix(result_df_wide_subset), show_rownames = F, fontsize = 8,
        color = viridis(100,option="B"), legend_breaks = c(0,20,40,60,80,100))

# Samples with high CE inclusion
## Subset to avgPSI > 10 for at least 42 targets (10% of all CEs)
## Subset to CEs within those samples with a max inclusion greater than 50
result_df_TDPneg <- result_df_wide_subset%>%
  mutate(testFilt = apply(result_df_wide_subset,
                     MARGIN = 1,
                     function(x) length(which(x > 10)))) %>%
  filter(testFilt > 42) %>%
  dplyr::select(-testFilt) 

colKeep <- apply(X = result_df_TDPneg, MARGIN = 2, FUN = function(x) max(x) > 50)
result_df_TDPneg <- result_df_TDPneg[,colKeep]

samples <- data.frame(sampleID = rownames(result_df_TDPneg))
sampleMeta <- snapmine::add_SnaptronMeta(samples, compilation = "sra_human") %>%
  mutate(rail_id = as.character(rail_id))
samples <- samples %>% full_join(sampleMeta, by = join_by("sampleID" == "rail_id")) %>%
  arrange(sample_title) %>% arrange(study_title) # Cluster related studies

# write.csv(samples, "/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Projects/NMD/NotGitHub/output_tables/251208highTDPSnapMine.csv")
#manual

# Sort samples manually
# Remove samples w/o duplicates (if sequenced) unless primary sample with another sample matching from a diff study
# Remove samples from CLIP-seq
samples_sorted <- read_excel("/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Projects/NMD/NotGitHub/output_tables/251208highTDPSnapMine_sorted.xlsx", sheet = 2)
result_df_TDPneg <- result_df_wide %>% filter(sampleID %in% samples_sorted$sampleID)

plotdf <- samples_sorted %>% dplyr::select(sampleID,sampleInfo, plottingGroup1) %>%
  left_join(result_df_TDPneg %>% mutate(sampleID = as.numeric(sampleID)), by = "sampleID")

rowAnn <- plotdf %>% dplyr::select(sampleInfo, plottingGroup1) %>% 
  mutate(sampleInfo = make_unique(sampleInfo, sep = " ")) %>% 
  column_to_rownames("sampleInfo")

plotdf <- plotdf %>% dplyr::select(-c(sampleID,plottingGroup1)) %>%
  mutate(sampleInfo = make_unique(sampleInfo, sep = " ")) %>% 
  column_to_rownames("sampleInfo") 

colKeep <- apply(X = plotdf, MARGIN = 2, FUN = function(x) median(x) > 10)
plotdf <- plotdf[,colKeep]

## Add CE values from i3N to plotting DF
ces_kept <- str_split_i(colnames(plotdf),"_",2)
ces_kept <- data.frame(ID = ces_kept) %>%
  left_join(df %>% dplyr::select(`ID #`,avgPSI_sgTARDBP, `maxPSI_sgTARDBP`),
            by = join_by(ID == `ID #`))

plotdf <- rbind(plotdf, data.frame(ces_kept$avgPSI_sgTARDBP)[,1])
plotdf <- rbind(plotdf, data.frame(ces_kept$maxPSI_sgTARDBP)[,1])
rownames(plotdf)[(nrow(plotdf)-1):nrow(plotdf)] = c("sgTDP-43", "sgTDP-43 + shUPF1/shXRN1")

rowAnn <- rbind(rowAnn, data.frame(plottingGroup1 = rep("TDP-43 KD",2)))
rownames(rowAnn)[(nrow(plotdf)-1):nrow(plotdf)] = c("sgTDP-43", "sgTDP-43 + shUPF1/shXRN1")

# Put i3N data at top
plotdf <- rbind(plotdf[(nrow(plotdf)-1):nrow(plotdf),], plotdf[1:(nrow(plotdf)-2),])

# Plot
pheatmap(as.matrix(plotdf), 
         show_rownames = T,
         cluster_rows = F,
         fontsize = 12,
         color = viridis(100,option="F"), 
         breaks = seq(0,100,1),
         border_color = NA,
         treeheight_col = 0,
         annotation_row = rowAnn)


```


## Compare PSIs of all targets to TDP-43 neg nuclei
Citation: Liu, Elaine Y *et al.* Loss of Nuclear TDP-43 Is Associated with Decondensation of LINE Retrotransposons. *Cell Rep* 27 1409-1421.e6(2019). doi:10.1016/j.celrep.2019.04.003

```{r TDPneg,fig.align = "c", out.width="100%", fig.dpi=300, fig.asp=1}
df <- fread("/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Scripts/NMDi_TDP43kd_repo/data/251126_CEdf.csv")
result_df_wide <- readRDS("/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Projects/NMD/NotGitHub/output_tables/251204_snapmine_result_df_wide.RDS")
Lee_sampleID <- fread("/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Projects/NMD/NotGitHub/data/SRP185789_Lee_SraRunTable.csv") %>%
  rename(sampleID = SnaptronID) %>% arrange(sampleID)%>% arrange(`genotype/variation`)

# Filter to Lee data
plot_df <- result_df_wide %>% 
  filter(sampleID %in% c(Lee_sampleID$sampleID)) 


## Add CE values from i3N to results
ces_kept <- str_split_i(colnames(result_df_wide),"_",2)
ces_kept <- data.frame("ID" = ces_kept) %>%
  left_join(df %>% dplyr::select(`ID #`,avgPSI_sgTARDBP, `maxPSI_sgTARDBP`),
            by = join_by(ID == `ID #`))

plot_df <- rbind(plot_df, ces_kept$avgPSI_sgTARDBP)
plot_df <- rbind(plot_df,ces_kept$`maxPSI_sgTARDBP`)

plot_df$sampleID[(nrow(plot_df)-1):nrow(plot_df)] <- c("sgTDP-43", "sgTDP-43 + NMDi")

# Change column to rownames for matrix
plot_df <- plot_df %>% column_to_rownames("sampleID") 

# Annotation
annotation <- data.frame(sampleID = c(Lee_sampleID$sampleID, "sgTDP-43", "sgTDP-43 + NMDi"),
                         sampleType = c(Lee_sampleID$`genotype/variation`, rep("i3N",2))) %>%
  column_to_rownames("sampleID")

# Set limits on values
plot_df[is.na(plot_df)] = 0

# Re-order
negIDs <- which(rownames(plot_df) %in% Lee_sampleID$sampleID[which(Lee_sampleID$`genotype/variation` == "TDP-43 negative")])
posIDs <- which(rownames(plot_df) %in% Lee_sampleID$sampleID[which(Lee_sampleID$`genotype/variation` == "TDP-43 positive")])

samples <- rbind(plot_df[negIDs,],plot_df[posIDs,])
plot_df <- rbind(plot_df[15:16,], samples)

plot_annotation <- annotation

# Which CEs have a higher PSI in at least one patient than in i3N without NMDi?
# TDPneg median - TDPpos median > 5
#  TDPneg max < 5
#  TDPpos max < 10
# TDP43 KD + NMDi max > 5
# (TDP43 KD + NMDi)-(TDP43 KD only) > 5

plotCEs <- which((apply(plot_df[3:9,], 2, median) - apply(plot_df[10:16,], 2, median) > 5) &
                   apply(plot_df[3:9,], 2, max) > 5 &
                   apply(plot_df[10:16,], 2, max) < 10 & apply(plot_df[2,], 2, max) > 5 &
                   apply(plot_df[2,], 2, median) - apply(plot_df[1,], 2, median) > 5)

                 
colOrder <- plotCEs[str_order((apply(plot_df[2,plotCEs], 2, median) - apply(plot_df[1,plotCEs], 2, median))/apply(plot_df[2,plotCEs], 2, median))]

### USE THIS PLOT
pheatmap(as.matrix(plot_df[,rev(colOrder)]), 
         show_colnames = T,
         cluster_cols = F,
         cluster_rows = F,
         fontsize = 12,
         color = viridis(100,option="F"), 
         breaks = seq(0,100,1),
         legend_breaks = c(0,50,100),
         border_color = NA,
         annotation_row = plot_annotation,
         display_numbers = F)


```



