---
title: "251220_lc"
author: "Irika Sinha"
date: "2025-12-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages

```{r packages, warning=FALSE}
library("BiocManager")
packages <- c("tidyverse", "data.table")

for(i in packages){
  if(!(i %in% rownames(installed.packages()))){
    BiocManager::install(pkgs = i, ask = F)
  } 
  library(package = i, character.only = T, quietly = T)
  rm(i)
}

rm(packages)

## List Package Versions
versions = installed.packages()[names(sessionInfo()$otherPkgs), "Version"]
data.frame(versions)
rm(versions)

```

## Load files

df: degrees of freedom, equal to the number of introns in the cluster minus one (assuming two groups)
groups should generally have 3 (cassette CE) introns, but with some additional differential splicing (slight shifts or messy splicing) might have up to 12 or 15? but this is only if there are many CEs within the same intron. df >= 2 if there is at least one alternate intron

```{r files}
setwd("/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Scripts/NMDi_TDP43kd_repo/")
cryptic_df <- fread("data/250601_CE_table.csv")

## Load LC output
sample_groups <- fread("data/Leafcutter/251220_sample_groups.csv")
knitr::kable(sample_groups)

## Load in all LC files
lc_df <- data.frame()

for(i in seq(1,9)){
  ## Load in both files and subset to significant events
  lc_df_sig <- fread(paste0("data/Leafcutter/group",i,"_cluster_significance.txt")) %>%
    filter(status == "Success")
  lc_df_sizes <- fread(paste0("data/Leafcutter/group",i,"_effect_sizes.txt")) %>%
    mutate(cluster = paste0(str_split_i(intron,":",1),":",str_split_i(intron,":",4))) 
  
  ## Merge files for each group
  colnames(lc_df_sizes)[3:4] <- c("group1", "group2")
  lc_df_sizes <- right_join(lc_df_sizes, lc_df_sig, by = "cluster") %>% 
    mutate(run = i,
           intronMod = paste0(str_split_i(intron,":",1),":",
                              str_split_i(intron,":",2),"-", 
                              str_split_i(intron,":",3))) %>% 
    dplyr::select(c(run, cluster, intronMod, group1, group2, df, deltapsi, p.adjust, genes))
  
  ## Add to overall df
  lc_df <- rbind(lc_df, lc_df_sizes)
  rm(lc_df_sig, lc_df_sizes,i)
}

# saveRDS(lc_df, "output_tables/251220_leafcutterResults.RDS")

```

## Create supplementary tables for paper - raw data from LC
Table S1: All significant events as a result of TDP-43 KD (Group 1 - Group 8 introns)
Table S2: All significant events as a result of TDP-43 KD only (Group 9 - Group 8 introns)
Table S3: All significant events identified through NMDi (Group 2 - Group 8 - Group 9 introns)
Table S4: All significant events identified through NMDi inhibiting XU (Group 3 - Group 8 - Group 9)

```{r supplementary}
lc_df <- readRDS("output_tables/251220_leafcutterResults.RDS") %>% filter(p.adjust < 0.05)

## Remove all introns significant in control vs NMDi samples (no TDP-43 KD)
lc_df <- lc_df %>% filter(!(intronMod %in% (lc_df %>% filter(run == 8))$intronMod)) %>% # Removes 52502 rows
  filter(group1 < 0.05, group2 > 0.1, !is.na(genes)) %>% # Filters for upper cut-off (g1) and lower cut-off (g2)
  mutate(p.adjust = signif(p.adjust,3))

## Table S1
lc_df_1 <- lc_df %>% filter(run == 1) %>%
  group_by(genes, cluster, p.adjust) %>%
  summarize(max_pctDiff = signif(max(deltapsi/(group1+0.01)),3),
            nIntrons = n(),
            introns = str_flatten_comma(intronMod),
            group1_PSIs = str_flatten_comma(signif(group1,3)),
            group2_PSIs = str_flatten_comma(signif(group2,3)),
            d_PSIs = str_flatten_comma(signif(deltapsi,3))) %>%
  dplyr::arrange(desc(max_pctDiff))

## Table S2
lc_df_2 <- lc_df %>% filter(run == 9) %>%
  group_by(genes, cluster, p.adjust) %>%
  summarize(max_pctDiff = signif(max(deltapsi/(group1+0.01)),3),
            nIntrons = n(),
            introns = str_flatten_comma(intronMod),
            group1_PSIs = str_flatten_comma(signif(group1,3)),
            group2_PSIs = str_flatten_comma(signif(group2,3)),
            d_PSIs = str_flatten_comma(signif(deltapsi,3))) %>%
  dplyr::arrange(desc(max_pctDiff))

## Table S3
lc_df_3 <- lc_df %>% filter(run == 2, 
                            !(intronMod %in% (lc_df %>% filter(run == 9))$intronMod)) %>% # Events not identified by TDP-43 kd only
  group_by(genes, cluster, p.adjust) %>%
  summarize(max_pctDiff = signif(max(deltapsi/(group1+0.01)),3),
            nIntrons = n(),
            introns = str_flatten_comma(intronMod),
            group1_PSIs = str_flatten_comma(signif(group1,3)),
            group2_PSIs = str_flatten_comma(signif(group2,3)),
            d_PSIs = str_flatten_comma(signif(deltapsi,3))) %>%
  dplyr::arrange(desc(max_pctDiff))

## Table S4
lc_df_4 <- lc_df %>% filter(run == 3, 
                            !(intronMod %in% (lc_df %>% filter(run == 9))$intronMod), # Events not identified by TDP-43 kd only
                            #(intronMod %in% (lc_df %>% filter(run == 2))$intronMod), # Events included in TDPkd + NMDi vs ctls
                            (intronMod %in% (lc_df %>% filter(run == 4))$intronMod)) %>% # Events included in TXU vs CXU
  group_by(genes, cluster, p.adjust) %>%
  summarize(max_pctDiff = signif(max(deltapsi/(group1+0.01)),3),
            nIntrons = n(),
            introns = str_flatten_comma(intronMod),
            group1_PSIs = str_flatten_comma(signif(group1,3)),
            group2_PSIs = str_flatten_comma(signif(group2,3)),
            d_PSIs = str_flatten_comma(signif(deltapsi,3))) %>%
  dplyr::arrange(desc(max_pctDiff))

# Save file
# writexl::write_xlsx(list(ST1 = lc_df_1, ST2 = lc_df_2, ST3 = lc_df_3, ST4 = lc_df_4), 
#                     path = "output_tables/251222_LC_ST.xlsx")


```

## Calculate p-value (adjusted) for identified CEs
```{r matchCE}
lc_df <- readRDS("output_tables/251220_leafcutterResults.RDS")

# Isolate junctions from identified CEs and re-format for LC
ce_junctions <- data.frame(junctions = unlist(str_split(c(cryptic_df$LeftJunction,cryptic_df$RightJunction),";"))) %>%
  filter(junctions != "NA") %>%
  mutate(chr = str_split_i(junctions, ":",1),
         coor = str_split_i(junctions, ":",2),
         start = as.numeric(str_split_i(coor, "-",1)),
         end = as.numeric(str_split_i(coor, "-",2)),
         lc_coord = paste0(chr,":", start-1, "-", end+1))

# Subset LC results to junctions of interest
lc_df_subset <- lc_df %>% filter(intronMod %in% ce_junctions$lc_coord)

# Create new dataframe and populate with p-values for TDPvsCtrland TXU+TXS vs TDP43 kd only
ce_df <- cryptic_df[,c(1:6,18,19)] %>% mutate(L_padj_TDPvCtrl = NA, R_padj_TDPvCtrl=NA,
                                              L_padj_TDPvTXUTXS = NA, R_padj_TDPvTXUTXS=NA)

for(i in 1:nrow(ce_df)){
  ## Left 1
  if(!is.na(ce_df$LeftJunction[i])){
    junctions <- unlist(str_split(ce_df$LeftJunction[i],";"))
    junctions <- ce_junctions$lc_coord[which(ce_junctions$junctions %in% junctions)]
    df <- lc_df_subset %>% filter(intronMod %in% junctions, run == 1)
  
    if(nrow(df) >= 1){
      df$p.adjust = signif(df$p.adjust, digits=3)
      ce_df$L_padj_TDPvCtrl[i] = str_flatten_comma(df$p.adjust)
    } else{
      ce_df$L_padj_TDPvCtrl[i] = "not calculated"
    }
    
    df <- lc_df_subset %>% filter(intronMod %in% junctions, run == 5)
    if(nrow(df) >= 1){
      df$p.adjust = signif(df$p.adjust, digits=3)
      ce_df$L_padj_TDPvTXUTXS[i] = str_flatten_comma(df$p.adjust)
    } else{
      ce_df$L_padj_TDPvTXUTXS[i] = "not calculated"
    }
  }
  
  
  ## Right 2
  if(!is.na(ce_df$RightJunction[i])){
    junctions <- unlist(str_split(ce_df$RightJunction[i],";"))
    junctions <- ce_junctions$lc_coord[which(ce_junctions$junctions %in% junctions)]
    df <- lc_df_subset %>% filter(intronMod %in% junctions, run == 1)
  
    if(nrow(df) >= 1){
      df$p.adjust = signif(df$p.adjust, digits=3)
      ce_df$R_padj_TDPvCtrl[i] = str_flatten_comma(df$p.adjust)
    } else{
      ce_df$R_padj_TDPvCtrl[i] = "not calculated"
    }
    
    df <- lc_df_subset %>% filter(intronMod %in% junctions, run == 5)
    if(nrow(df) >= 1){
      df$p.adjust = signif(df$p.adjust, digits=3)
      ce_df$R_padj_TDPvTXUTXS[i] = str_flatten_comma(df$p.adjust)
    }  else{
      ce_df$R_padj_TDPvTXUTXS[i] = "not calculated"
    }
  }
  
  rm(i, df)
}

# write.csv(ce_df,"output_tables/251221_LC_sig.csv")


```


## Find un-identified events (mainly skiptic)

```{r filtersNew}
lc_df <- readRDS("output_tables/251220_leafcutterResults.RDS") %>% 
  filter(p.adjust < 0.05,
         group1 < 0.05, group2 > 0.1,  df > 1, !is.na(genes),
         (run %in% c(1,3, 4,9))) %>% # Filters for upper cut-off (g1) and lower cut-off (g2)
  mutate(p.adjust = signif(p.adjust,3))

lc_df <- lc_df  %>% filter(!(intronMod %in% (lc_df %>% filter(run == 8))$intronMod),
                           !(intronMod %in% lc_df_subset$intronMod)) # Removes identified

lc_df_tabulate <- lc_df  %>% 
  group_by(genes, cluster) %>%
  summarize(runs = str_flatten_comma(c(run)),
            maxdPSI = signif(max(deltapsi),3),
            max_pctDiff = signif(max(deltapsi/(group1+0.01)),3),
            nIntrons = n(),
            introns = str_flatten_comma(intronMod),
            group1_PSIs = str_flatten_comma(signif(group1,3)),
            group2_PSIs = str_flatten_comma(signif(group2,3)),
            d_PSIs = str_flatten_comma(signif(deltapsi,3))) %>%
  dplyr::arrange(desc(max_pctDiff)) %>%
  filter(runs != "3",maxdPSI > 0.1,
         ifelse(str_detect(runs,"9"), str_detect(runs,"(4|7)+"),str_detect(runs,"1|3|4")), # If detected in TDP-43 only, should be detected in NMDi conditions
         ifelse(str_detect(runs,"4"), str_detect(runs,"3"),str_detect(runs,"1"))) # I

 # write.csv(lc_df_tabulate,"output_tables/251221_LC_new.csv")



```



