---
title: "251220_lc"
author: "Irika Sinha"
date: "2025-12-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages

```{r packages, warning=FALSE}
library("BiocManager")
packages <- c("tidyverse", "data.table")

for(i in packages){
  if(!(i %in% rownames(installed.packages()))){
    BiocManager::install(pkgs = i, ask = F)
  } 
  library(package = i, character.only = T, quietly = T)
  rm(i)
}

rm(packages)

## List Package Versions
versions = installed.packages()[names(sessionInfo()$otherPkgs), "Version"]
data.frame(versions)
rm(versions)

```

## Load files

df: degrees of freedom, equal to the number of introns in the cluster minus one (assuming two groups)
groups should generally have 3 (cassette CE) introns, but with some additional differential splicing (slight shifts or messy splicing) might have up to 12 or 15? but this is only if there are many CEs within the same intron. df >= 2 if there is at least one alternate intron

```{r files}
setwd("/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Scripts/NMDi_TDP43kd_repo/")
cryptic_df <- fread("data/250601_CE_table.csv")

## Load LC output
sample_groups <- fread("data/Leafcutter/251220_sample_groups.csv")
knitr::kable(sample_groups)

## Load in all LC files
lc_df <- data.frame()

for(i in seq(1,9)){
  ## Load in both files and subset to significant events
  lc_df_sig <- fread(paste0("data/Leafcutter/group",i,"_cluster_significance.txt")) %>%
    filter(status == "Success")
  lc_df_sizes <- fread(paste0("data/Leafcutter/group",i,"_effect_sizes.txt")) %>%
    mutate(cluster = paste0(str_split_i(intron,":",1),":",str_split_i(intron,":",4))) 
  
  ## Merge files for each group
  colnames(lc_df_sizes)[3:4] <- c("group1", "group2")
  lc_df_sizes <- right_join(lc_df_sizes, lc_df_sig, by = "cluster") %>% 
    mutate(run = i,
           intronMod = paste0(str_split_i(intron,":",1),":",
                              str_split_i(intron,":",2),"-", 
                              str_split_i(intron,":",3))) %>% 
    dplyr::select(c(run, cluster, intronMod, group1, group2, df, deltapsi, p.adjust, genes))
  
  ## Add to overall df
  lc_df <- rbind(lc_df, lc_df_sizes)
  rm(lc_df_sig, lc_df_sizes,i)
}

# saveRDS(lc_df, "output_tables/251220_leafcutterResults.RDS")

```

## ID'ed CE padj
```{r matchCE}
ce_junctions <- data.frame(junctions = unlist(str_split(c(cryptic_df$LeftJunction,cryptic_df$RightJunction),";"))) %>%
  filter(junctions != "NA") %>%
  mutate(chr = str_split_i(junctions, ":",1),
         coor = str_split_i(junctions, ":",2),
         start = as.numeric(str_split_i(coor, "-",1)),
         end = as.numeric(str_split_i(coor, "-",2)),
         lc_coord = paste0(chr,":", start-1, "-", end+1))

lc_df_subset <- lc_df %>% filter(intronMod %in% ce_junctions$lc_coord)

ce_df <- cryptic_df[,c(1:6,18,19)] %>% mutate(L_padj_TDPvCtrl = NA, R_padj_TDPvCtrl=NA,
                                              L_padj_TDPvTXUTXS = NA, R_padj_TDPvTXUTXS=NA)

for(i in 1:nrow(ce_df)){
  ## Left 1
  if(!is.na(ce_df$LeftJunction[i])){
    junctions <- unlist(str_split(ce_df$LeftJunction[i],";"))
    junctions <- ce_junctions$lc_coord[which(ce_junctions$junctions %in% junctions)]
    df <- lc_df_subset %>% filter(intronMod %in% junctions, run == 1)
  
    if(nrow(df) >= 1){
      df$p.adjust = signif(df$p.adjust, digits=3)
      ce_df$L_padj_TDPvCtrl[i] = str_flatten_comma(df$p.adjust)
    } else{
      ce_df$L_padj_TDPvCtrl[i] = "not calculated"
    }
    
    df <- lc_df_subset %>% filter(intronMod %in% junctions, run == 5)
    if(nrow(df) >= 1){
      df$p.adjust = signif(df$p.adjust, digits=3)
      ce_df$L_padj_TDPvTXUTXS[i] = str_flatten_comma(df$p.adjust)
    } else{
      ce_df$L_padj_TDPvTXUTXS[i] = "not calculated"
    }
  }
  
  
  ## Right 2
  if(!is.na(ce_df$RightJunction[i])){
    junctions <- unlist(str_split(ce_df$RightJunction[i],";"))
    junctions <- ce_junctions$lc_coord[which(ce_junctions$junctions %in% junctions)]
    df <- lc_df_subset %>% filter(intronMod %in% junctions, run == 1)
  
    if(nrow(df) >= 1){
      df$p.adjust = signif(df$p.adjust, digits=3)
      ce_df$R_padj_TDPvCtrl[i] = str_flatten_comma(df$p.adjust)
    } else{
      ce_df$R_padj_TDPvCtrl[i] = "not calculated"
    }
    
    df <- lc_df_subset %>% filter(intronMod %in% junctions, run == 5)
    if(nrow(df) >= 1){
      df$p.adjust = signif(df$p.adjust, digits=3)
      ce_df$R_padj_TDPvTXUTXS[i] = str_flatten_comma(df$p.adjust)
    }  else{
      ce_df$R_padj_TDPvTXUTXS[i] = "not calculated"
    }
  }
  
  rm(i, df)
}

# write.csv(ce_df,"output_tables/251221_LC_sig.csv")


```


## Find un-identified events (mainly skiptic)
1) Main groups will be group 1, group
```{r filtersNew}
lc_df <- readRDS("output_tables/251220_leafcutterResults.RDS") %>% 
  filter(p.adjust < 0.05 & df > 1,
         deltapsi/group1 > 0.3, str_detect(genes,",", negate=T),
        !(run %in% c(2,3,5,6))) # g2 expression should be at least 30% more

ctrlIntrons <- unique((lc_df %>% filter(run == 8))$intronMod)

# Look for any missed CEs
lc_df_tabulate <- lc_df  %>%
  filter(ifelse(run %in% c(1,4,7,9), group1 < 0.10, group2/group1 > 0.2 ),
         run != 7) %>%
  group_by(intronMod, genes) %>%
  summarize(runs = str_flatten_comma(c(run)), 
            maxdPSI = round(max(deltapsi),2), # Identify max dPSI between g1 and g2 for each intron
            max1PSI = max(group1), # Identify max PSI of g1 for each intron
            max2PSI = round(max(group2),2)) %>% # Identify max PSI of g2 for each intron
  filter(str_detect(runs,","), # Should be identified in multiple groupings
         str_detect(runs,"1"), # Should be identified in all KD vs all ctrl
         ifelse(str_detect(runs,"9"), str_detect(runs,"(4|7)+"),str_detect(runs,"4|7")),
         maxdPSI > 0.01, # g1 vs g2 PSIs should be at least 1% different
         max2PSI > 0.2, # g2 PSI should be greater than 1%
         !(intronMod %in% ctrlIntrons)
         ) 

new_df <- lc_df_tabulate %>% filter(!(intronMod %in% lc_df_subset$intronMod)) %>%
  mutate(order = maxdPSI/(max1PSI + 0.01) )

# write.csv(new_df,"output_tables/251221_LC_new.csv")



```



