---
title: "251117_qPCR"
author: "Irika Sinha"
date: "2025-11-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages, message=FALSE}
library("BiocManager")
packages <- c("tidyverse", "ggrepel","viridis", "ggridges", "pheatmap","readxl")

for(i in packages){
  if(!(i %in% rownames(installed.packages()))){
    BiocManager::install(pkgs = i, ask = F)
  } else{
    #BiocManager::install(pkgs = i, ask = F, update=T)
  }
  library(package = i, character.only = T, quietly = T)
  rm(i)
}

rm(packages)

## List Package Versions
versions = installed.packages()[names(sessionInfo()$otherPkgs), "Version"]
data.frame(versions)
rm(versions)

```

## Data QC
<<<<<<< Updated upstream
11/17/2025
-KALRN #7-21 & #7-22 were not amplified (no primer added)
-Primers to potentially re-evaluate: AKT3, ATG4B, CBARP, INTS8, *ITGA7*, MADD, *MAST1*, *PHF2*, PKN1, NSMCE4A, RAP1GAP, ZNF23

```{r DataQC}
setwd("/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Scripts/NMDi_TDP43kd_repo/")
data <- read_csv("data/251117_qPCR_data.csv") %>% 
  filter(!is.na(Primers)  & Primers != "NA" & CT != "Undetermined") %>%
  mutate(UI = paste0(Target,"_", SampleID,"_",Primers))
data_2 <- readxl::read_xlsx("data/251218_qPCR_Calculations.xlsx") %>% 
  filter(!is.na(Primers) & Primers != "NA" & CT != "Undetermined") %>%
  mutate(Primers = ifelse(Primers == Target, "Target", Primers),
         UI = paste0(Target,"_", SampleID,"_",Primers))
=======

```{r DataQC}
setwd("/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Scripts/NMDi_TDP43kd_repo/")
# setwd("/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Projects/NMD/NotGitHub")
data <- read_xlsx("data/data_qPCR_all.xlsx", sheet=1) %>% 
  filter(!is.na(Primers)  & Primers != "NA" & CT != "Undetermined" & EXPFAIL != "Y") %>%
  mutate(Primers = ifelse(Target == Primers, "Target", Primers),
         UI = paste0(Target,"_", SampleID,"_",Primers),
         Target = ifelse(SampleID == "TARDBP_KD", "TARDBP", Target))
>>>>>>> Stashed changes
shRNA_key <- read_xlsx("data/250905_qPCR_RNAs.xlsx", sheet = 1) %>% 
  dplyr::select(gene, id, shRNA) %>% distinct() 
batches = rev(unique(data$Plate))


## Only keep used shRNA
shRNA_keep <- readxl::read_xlsx("/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Papers/NMD/02_Revisions/TDP43-NMD_Tables_Supplementary.xlsx", sheet = 8) %>%
  filter(`TRC Number` != "N/A")
shRNA_keep <- shRNA_keep$`TRC Number`

for(i in 1:nrow(shRNA_key)) {
  trcn <- str_split_i(shRNA_key$shRNA[i], "TRCN", 2) 

  if(!is.na(trcn)){
    trcn <- str_which(shRNA_keep, trcn)
    shRNA_key$shRNA[i] = shRNA_keep[trcn]
  }
}

shRNA_key <- shRNA_key %>% filter(shRNA %in% c(shRNA_keep, "shCTR", "shTDP-43"))

## Calculate average CT and sd of CT
<<<<<<< Updated upstream
data <- data %>% group_by(Target, SampleID, Primers,UI) %>% 
=======
data_summ <- data %>% group_by(Plate,Target, SampleID, Primers,UI) %>% 
>>>>>>> Stashed changes
  summarize(ct1 = round(as.numeric(list(CT)[1][[1]][1]),2), ct2=round(as.numeric(list(CT)[1][[1]][2]),2),
            ct_diff=abs(ct2-ct1),
            avgCT = round(mean(c(ct2,ct1)),2), 
            sdCT = round(sd(c(ct2,ct1)),2)) %>%
  filter(!is.na(ct_diff))

<<<<<<< Updated upstream
data_2 <- data_2 %>% group_by(Target, SampleID, Primers,UI) %>% 
  summarize(ct1 = round(as.numeric(list(CT)[1][[1]][1]),2), ct2=round(as.numeric(list(CT)[1][[1]][2]),2),
            ct_diff=abs(ct2-ct1),
            avgCT = round(mean(c(ct2,ct1)),2), 
            sdCT = round(sd(c(ct2,ct1)),2))

## Replace samples that are re-done in data df and then merge
# Data #2 includes samples re-done due to Ct differences > 0.5
df <- data.frame()
df_2 <- data_2

for(i in 1:nrow(data)){
  # Was this experiment re-done?
  if(data$UI[i] %in% df_2$UI){
    row = which(df_2$UI == data$UI[i]) # Identify row for new experiment
    df <- rbind(df, df_2[row,]) # Add new data
    df_2 <- df_2[-row,] # remove from df_2
    rm(row)
  } else{
    # Experiment was not re-done
    df <- rbind(df, data[i,]) # Add original data
  }
  rm(i)
}

df <- rbind(df, df_2)

# Does not pass - check against first two values
potential_redo <- df %>% filter(ct_diff > 0.5)
prev_redo <- potential_redo[which(potential_redo$UI %in% data$UI),] %>%
  left_join(data, by = c("Target", "SampleID","Primers","UI"), suffix = c("_2","_1")) %>%
  rowwise() %>%
  mutate(avg_overall = mean(c(ct1_1, ct1_2, ct2_1, ct2_2)),
         sd_overall = sd(c(ct1_1, ct1_2, ct2_1, ct2_2)))

## Check all
=======
## Replace samples that are re-done in data df and then merge
df <- data.frame()
targets <- unique(data$UI)

for(i in targets){
  df_temp <- data_summ %>% filter(UI == i)
  rows <- which(df_temp$ct_diff == min(df_temp$ct_diff))
  
  df <- rbind(df, df_temp[rows,])
  
  rm(df_temp, rows, i)
}
>>>>>>> Stashed changes

## Check all
df %>% mutate(key = paste0(Target, "_", SampleID)) %>%
  filter(Primers %in% c("Target", "TARDBP", "GAPDH"))%>%
  ggplot(aes(x=key, y = ct_diff, fill = Primers))+
  geom_col(position = "dodge")+
  scale_fill_viridis(option = "B", discrete=T, begin = 0, end = 0.8, name="Primers")+
  xlab("KDTarget_shRNA")+theme_classic() + ylab("CT Difference for Duplicates")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  scale_x_discrete(expand=c(0,0)) + 
  scale_y_continuous(expand=c(0,0), lim=c(0,2)) +
  geom_hline(yintercept = 0.5, linetype="dotted") 
  labs(title = "All experiments")

df %>% filter(SampleID %in% c("Ctrl", "TARDBP_KD"))%>%
  ggplot(aes(x=Primers, y = ct_diff, fill = SampleID))+
  geom_col(position = "dodge")+
  scale_fill_viridis(option = "B", discrete=T, begin = 0, end = 0.8, name="Sample")+
  xlab("Primers")+theme_classic() + ylab("CT Difference for Duplicates")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  scale_x_discrete(expand=c(0,0)) + 
  scale_y_continuous(expand=c(0,0), lim=c(0,1)) +
  geom_hline(yintercept = 0.5, linetype="dotted") 
  labs(title = "Ctrl and TDP-43 KD samples")
  
df %>% mutate(key = paste0(Target, "_", SampleID),
              Primers = ifelse(!(Primers %in% c("Target", "TARDBP", "GAPDH")), 
                               "Target", Primers)) %>%
  ggplot(aes(x=ct1, y = ct2, color = Primers))+
  geom_point(size=3, alpha=0.8)+theme_classic()+
  scale_color_viridis(option = "B", discrete=T, begin = 0, end = 0.8, name="Primers")+
  geom_abline(slope=1, intercept=-0.5, linetype="dotted")+
  geom_abline(slope=1, intercept=0.5, linetype="dotted")+
  geom_text_repel(aes(label=ifelse(ct_diff>0.5,SampleID,"")),
                  size=3,min.segment.length = 0,
                  max.time=2,max.overlaps=20, nudge_y = 0.1,  direction="both")

<<<<<<< Updated upstream
## Re-do samples with CT difference > 0.5
potentialRedos <- (df %>% filter(ct_diff > 0.5)) %>% arrange(SampleID) %>% arrange(Primers)
#write.csv(potentialRedos,"output_tables/251218_qPCR_redo.csv")


```



```{r ddct}
## Add shRNA metadata
df <- df %>% left_join(shRNA_key, by = join_by(SampleID == id, Target == gene)) %>% 
  dplyr::select(-UI)

## Calculate -ddCT for okay samples
calculate_df_Ctrl <- data_grouped %>% 
  filter(SampleID == "Ctrl") %>% ungroup() %>%
  mutate(GAPDH = data_grouped$avgCT[480],
         TARDBP = data_grouped$avgCT[497],
         Ctrl_dCT_Target = as.numeric(avgCT)-as.numeric(GAPDH),
         Ctrl_dCT_TDP = as.numeric(TARDBP)-as.numeric(GAPDH)) %>%
  dplyr::select(!c(Target,shRNA, ct1,ct2,ct_diff,sdCT, avgCT, GAPDH, TARDBP, SampleID))

calculate_df_TDP43KD <- data_grouped %>% 
  filter(SampleID == "TARDBP_KD" & Primers %in% unique(data_grouped$Target[-which(data_grouped$SampleID %in% potentialRedos)])[-43]) %>% ungroup() %>%
  left_join(calculate_df_Ctrl, by = "Primers") %>%
    mutate(GAPDH = data_grouped$avgCT[which(data_grouped$SampleID == "TARDBP_KD" & data_grouped$Primers == "GAPDH")],
         TDPKD_dCT_Target = as.numeric(avgCT)-as.numeric(GAPDH),
         ddCT_TDPKD_Target = TDPKD_dCT_Target - Ctrl_dCT_Target,
         relTarget_TDPKD = round(2^(-1*ddCT_TDPKD_Target),3)) 

calculate_df_kds <- data_grouped %>% 
  filter(!(SampleID %in% c(potentialRedos,"TARDBP_KD", "Ctrl"))) %>%
  rename(KD=Target) %>%
  dplyr::select(!c(ct1,ct2,ct_diff,sdCT)) %>%
  pivot_wider(names_from = (Primers), values_from = avgCT) %>%
  left_join(calculate_df_Ctrl, by = join_by(KD == Primers)) %>%
    mutate(dCT_Target = as.numeric(Target)-as.numeric(GAPDH),
         dCT_TDP = as.numeric(TARDBP)-as.numeric(GAPDH),
         ddCT_Target = dCT_Target - Ctrl_dCT_Target,
         ddCT_TARDBP = dCT_TDP - Ctrl_dCT_TDP,
         relTarget = round(2^(-1*ddCT_Target),3),
         relTDP43 = round(2^(-1*ddCT_TARDBP),3)) %>%
  ungroup() %>%
  add_row(KD = calculate_df_TDP43KD$Primers, shRNA = "shTARDBP",
          Ctrl_dCT_Target = calculate_df_TDP43KD$Ctrl_dCT_Target,
          dCT_Target = calculate_df_TDP43KD$TDPKD_dCT_Target,
          ddCT_Target = calculate_df_TDP43KD$ddCT_TDPKD_Target,
          relTarget = calculate_df_TDP43KD$relTarget_TDPKD) %>%
  mutate(color = ifelse(shRNA == "shTARDBP", "red", "black"))

tdpKDintercept <- data.frame(KD = calculate_df_TDP43KD$Primers, Val = calculate_df_TDP43KD$relTarget_TDPKD)

calculate_df_kds %>% 
  mutate(key = paste0(KD, "\n", shRNA)) %>%
  filter(!is.na(Ctrl_dCT_Target))%>%
  ggplot(aes(x = key, y = relTarget)) +
  geom_hline(data = tdpKDintercept, aes(yintercept = Val), linetype="dotted", color = "red", linewidth=1)+
  geom_boxplot(outliers = F, alpha=0.5)+
  geom_jitter(aes(color = color), size = 3, alpha = 0.6,height=0)+
  theme_classic() + scale_color_manual(values = c("black", "red")) + 
  scale_y_continuous(lim=c(0,2.4), expand=c(0,0))+
  xlab("")+ylab("Relative Expression of Target (2^-ddCT)")+
  facet_wrap(vars(KD), scales = "free_x", ncol=13)+
=======

```

## Compare log2FC of gene after shRNA vs TDP-43 KD
dCt = Ct(target gene) - Ct(ref gene)
ddCT = dCt (kd) - dCT (ctrl)

ddCT (shRNA log2FC)
ddCT(KD log2FC) for each gene

dCT ctrl = Ct(target) - Ct(ref gene) when target is NA



```{r ddct}
## Add shRNA metadata
df_info <- df %>% left_join(shRNA_key, by = join_by(SampleID == id, Target == gene)) %>% 
  ungroup() %>%
  dplyr::select(-c(Plate))  %>%
  mutate(shRNA = ifelse(Target == "NA", "shCTR", ifelse(Target == "TARDBP", "shTDP", shRNA)))

## Calculate dCT
df_calc_shRNA <- shRNA_key %>% filter(!(gene %in% c("Ctrl","TARDBP"))) %>%
  left_join(df_info %>% filter(!(Target %in% c("NA", "TARDBP")), Primers == "Target") %>% 
              dplyr::select(Target, SampleID, avgCT) %>% 
              rename(CT_TARGET_shRNA = "avgCT"),
              by = join_by("id" == "SampleID", "gene" == "Target")) %>%
    left_join(df_info %>% filter(!(Target %in% c("NA", "TARDBP")), Primers == "GAPDH") %>% 
                dplyr::select(Target, SampleID, avgCT) %>% 
                rename(CT_REF_shRNA = "avgCT"),
              by = join_by("id" == "SampleID", "gene" == "Target")) %>%
  mutate(dCT_shRNA = CT_TARGET_shRNA - CT_REF_shRNA)

df_calc_ctrl <- shRNA_key %>% filter(!(gene %in% c("Ctrl","TARDBP"))) %>%
  dplyr::select("gene") %>% unique() %>%
  left_join(df_info %>% filter(Target == "NA") %>% 
              dplyr::select(Primers, avgCT) %>% 
              rename(CT_TARGET_ctrl = "avgCT"),
            by = join_by("gene" == "Primers")) %>%
  mutate(CT_REF_ctrl = df_info$avgCT[which(df_info$Target == "NA" & df_info$Primers == "GAPDH")],
         dCT_ctrl = CT_TARGET_ctrl - CT_REF_ctrl)

df_calc_shTDP <- shRNA_key %>% filter(!(gene %in% c("Ctrl","TARDBP"))) %>%
  dplyr::select("gene") %>% unique() %>%
  left_join(df_info %>% filter(Target == "TARDBP") %>% 
              dplyr::select(Primers, avgCT) %>% 
              rename(CT_TARGET_shTDP = "avgCT"),
            by = join_by("gene" == "Primers")) %>%
  mutate(CT_REF_shTDP = df_info$avgCT[which(df_info$Target == "TARDBP" & df_info$Primers == "GAPDH")],
         dCT_shTDP = CT_TARGET_shTDP - CT_REF_shTDP)

## Calculate ddCT
df_calc_shRNA <- df_calc_shRNA %>% dplyr::select(gene, id, shRNA, dCT_shRNA) %>%
  left_join(df_calc_ctrl %>% dplyr::select(gene, dCT_ctrl),
            by = "gene") %>%
  mutate(ddCT_shRNA = round(dCT_shRNA - dCT_ctrl,2), 
         fc = round(2^(ddCT_shRNA * -1),4))
df_calc_shTDP <- df_calc_shTDP %>% dplyr::select(gene, dCT_shTDP) %>%
  left_join(df_calc_ctrl %>% dplyr::select(gene, dCT_ctrl),
            by = "gene") %>%
  mutate(ddCT_shTDP = round(dCT_shTDP - dCT_ctrl,2), 
         fc = round(2^(-ddCT_shTDP),4))
df_calc <- rbind(df_calc_shRNA %>% dplyr::select(gene, shRNA, fc),
                 df_calc_shTDP %>% mutate(shRNA = "shTDP") %>% dplyr::select(gene, shRNA, fc))

tdpKDintercept <-  data.frame(gene = df_calc$gene) %>%
  left_join(df_calc_shTDP, by = "gene") %>%
  dplyr::select(gene, fc) %>%
  distinct()

## PLOT LOG2FC w TDP comparison
df_calc %>%  mutate(col = ifelse(shRNA == "shTDP", "red", "black"),
                    fc = ifelse(fc > 2, 2, fc)) %>%
  ggplot(aes(x = shRNA, y = fc)) +
  geom_hline(data = tdpKDintercept, aes(yintercept = fc), linetype="dotted", color = "red", linewidth=1)+
  geom_boxplot(outliers = F, alpha=0.5) +
  geom_jitter(aes(color = col), size = 2, alpha = 0.6,height=0)+
  theme_classic() + scale_color_manual(values = c("black", "red")) +
  scale_y_continuous(limits = c(0,2.1), n.breaks = 5,breaks = c(0,0.5,1, 1.5, 2), labels = c(0,0.5,1, 1.5,">=2"))+
  ylab("") +
  facet_wrap(vars(gene), scales = "free_x", ncol=14) +
>>>>>>> Stashed changes
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1),
        panel.border = element_rect(color="black"),
        strip.background = element_rect(fill="#B0DAF1", color="#B0DAF1"),
        strip.text = element_text(face="bold"),
        text = element_text(size =15),
        strip.text.x = element_text(size=12),
        legend.position = "none")

## PLOT LOG2FC with TDP comparison (summary)
df_calc %>% mutate(shRNA = ifelse(!(shRNA %in% c("NA", "shTDP")), "Target", shRNA),
                   fc = ifelse(fc > 1, 1, fc)) %>%
  group_by(gene, shRNA) %>% summarize(fc = median(fc))  %>%
  pivot_wider(names_from = shRNA, values_from = fc) %>%
  ggplot(aes(x = as.numeric(Target), y = as.numeric(shTDP))) +
  geom_jitter(size = 4, alpha = 0.6,height=0)+
  theme_classic() + geom_abline(intercept = 0, slope = 1, color = "#820000", linetype = "dashed") +
  xlim(0,1) + ylim(0,1)+
  # scale_y_continuous(limits = c(0,2.1), n.breaks = 5,breaks = c(0,0.5,1, 1.5, 2), labels = c(0,0.5,1, 1.5,">=2"))+
  ylab("shTDP GEX FC") +xlab("shTarget GEX FC") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1),
        text = element_text(size =15),
        legend.position = "none")


## Calculate ddCT of TDP-43 in all conditions
df_calc_TDP <- df_info %>% filter(Primers == "TARDBP") %>%
  left_join(df_info %>% filter(!(Target %in% c("NA", "TARDBP")), Primers == "TARDBP") %>% 
              dplyr::select(Target, SampleID, avgCT) %>% 
              rename(CT_TDP_shRNA = "avgCT"),
              by = join_by(SampleID, Target)) %>%
  left_join(df_info %>% filter(!(Target %in% c("NA", "TARDBP")), Primers == "GAPDH") %>% 
                dplyr::select(Target, SampleID, avgCT) %>% 
                rename(CT_REF_shRNA = "avgCT"),
              by = join_by(SampleID, Target)) %>%
  dplyr::select(-avgCT) %>%
  rbind(df_info %>% filter(Target == "TARDBP", Primers == "TARDBP") %>%
        rename(CT_TDP_shRNA = "avgCT") %>%
        mutate(CT_REF_shRNA =  (df_info %>% filter(Target == "TARDBP", Primers == "GAPDH"))$avgCT)) %>% 
  mutate( CT_TDP_ctrl = (df_info %>% filter(Target == "NA", Primers == "TARDBP") %>% 
              dplyr::select(SampleID, avgCT))$avgCT) %>%
  mutate( CT_REF_ctrl = (df_info %>% filter(Target == "NA", Primers == "GAPDH") %>% 
              dplyr::select(SampleID, avgCT))$avgCT) %>%
  mutate(dCT_TDP_shRNA = CT_TDP_shRNA - CT_REF_shRNA,
         dCT_TDP_Ctrl = CT_TDP_ctrl - CT_REF_ctrl,
         ddCT = dCT_TDP_shRNA - dCT_TDP_Ctrl,
         fc = 2^(-ddCT),
         key = paste0(Target, ": ",shRNA),
         key = ifelse(shRNA == "shTDP", "shTDP-43", ifelse(shRNA == "shCTR", "Ctrl", key)),
         key = str_remove_all(key,"0")) %>%
  filter(!is.na(fc)) 

shRNA_plot <- df_calc_TDP %>% group_by(Target, shRNA) %>% summarise(fc = mean(fc))
shRNA_dup <- shRNA_plot$Target[which(duplicated(shRNA_plot$Target))]

for(i in shRNA_dup) {
  df_temp <- shRNA_plot %>% filter(Target == i)
  trc <- df_temp$shRNA[which(df_temp$fc == max(df_temp$fc))]
  shRNA_plot <-  shRNA_plot %>% filter(shRNA != trc)
}

order_tdpshRNA <- (df_calc_TDP %>% filter(shRNA %in% c(shRNA_plot$shRNA, "shTDP")) %>% 
                     group_by(Target) %>% summarize(orderMean = mean(fc, na.rm=T), ordersd = sd(fc)) %>% filter(orderMean < 1, ordersd < 0.1 | is.na(ordersd)) %>% arrange(orderMean))$Target

## Plot TDP-43 RNA levels in single target knockdowns with sdFC < 0.1
df_calc_TDP %>% filter(Target %in% order_tdpshRNA & shRNA %in% c(shRNA_plot$shRNA, "shTDP"))%>%
  mutate( Target = factor(Target, order_tdpshRNA)) %>%
  ggplot(aes(x = Target, y = fc, fill=fc)) +
  geom_boxplot(outliers = F, alpha=0.8) +
  geom_jitter(size = 2, alpha = 0.6,height=0)+
  theme_classic() +
  scale_fill_gradient(low = "#820000", high="white", lim = c(0,1))+
  scale_y_continuous(limits = c(0,1.05), n.breaks = 4,breaks = c(0,0.25,0.5,0.75, 1), expand = c(0,0))+
  ylab("") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1),
        text = element_text(size =15))

df_plot <- df_calc_TDP %>%
  filter(Target %in% order_tdpshRNA &
         shRNA %in% c(shRNA_plot$shRNA, "shTDP")) %>%
  group_by(Target) %>%
  mutate(fc_group = median(fc, na.rm = TRUE)) %>%   
  ungroup() %>%
  mutate(Target = factor(Target, order_tdpshRNA))

ggplot(df_plot, aes(x = Target, y = fc, fill = fc_group)) +
  geom_boxplot(outliers = FALSE, alpha = 0.8) +
  geom_jitter(size = 2, alpha = 0.6, height = 0) +
  theme_classic() +
  scale_fill_gradient(low = "#820000", high = "white", limits = c(0,1)) +
  scale_y_continuous(limits = c(0,1.1),
                     breaks = c(0,0.25,0.5,0.75,1),
                     expand = c(0,0)) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        text = element_text(size = 15))

```