---
title: "260330_Proteomics_"
author: "Irika Sinha"
date: "2026-03-31"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Install packages

```{r packages, message=FALSE, warning=FALSE}
library("BiocManager")
packages <- c("tidyverse","viridis", "ggplot2", "readxl", "UpSetR","gghighlight")

for(i in packages){
  if(!(i %in% rownames(installed.packages()))){
    BiocManager::install(pkgs = i, ask = F, quietly = T)
  } else if (!(BiocManager::valid(i))){
    BiocManager::install(pkgs = i, ask = F, update = T, quietly = T)
  }
  library(package = i, character.only = T, quietly = T)
  rm(i)
}

rm(packages)

## List Package Versions
versions = installed.packages()[names(sessionInfo()$otherPkgs), "Version"]
data.frame(versions)
rm(versions)

```
## Data from i3N and patient proteomics
Integrative multiomic analysis links TDP-43-driven splicing defects to cascading proteomic disruption of ALS/FTD pathways
Velina Kozareva, Zhengde Liu, Keyana Blake, Yue A. Qi, Sasha Rollinson, Sahba Seddighi, Mohammad Alsaidi, Stanislav Tsitkov, Mercedes Prudencio, Leonard Petrucelli, Dennis W. Dickson, Anna-Leigh Brown, Pietro Fratta, Hong Joo Kim, J. Paul Taylor, Michael E. Ward, Ernest Fraenkel, Sarah E. Kargbo-Hill
bioRxiv 2025.09.29.679403; doi: https://doi.org/10.1101/2025.09.29.679403


```{r data}
df_deps <- readxl::read_xlsx('/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Projects/NMD/NotGitHub/data/Kozareva_proteomics.xlsx', sheet=1)

## cryptic and skiptic exons
cryptic_df <- readxl::read_xlsx("/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Papers/NMD/02_Revisions/260331_SharedCopies/TDP43-NMD_Tables_Main.xlsx", sheet = 2) %>% 
  dplyr::select(gene_symbol,ensembl_id, maxPSI_Ctrl, avgPSI_sgTARDBP, maxPSI_sgTARDBP)                                                                                                                                               
skiptic_df <- readxl::read_xlsx("/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Papers/NMD/02_Revisions/260331_SharedCopies/TDP43-NMD_Tables_Main.xlsx", sheet = 3) %>% 
  dplyr::select(gene_symbol,ensembl_id, maxPSI_Ctrl, avgPSI_sgTARDBP, maxPSI_sgTARDBP) 

df_ces <- rbind(cryptic_df,skiptic_df) %>% 
  mutate(gene_NMD = ifelse(avgPSI_sgTARDBP < 10,
                           ifelse(avgPSI_sgTARDBP < 5, "5pct", "10pct"),
                           "TDP-43 KD only")) %>%
  group_by(gene_symbol, ensembl_id) %>% summarise(psi_max = max(maxPSI_sgTARDBP), gene_NMD = str_flatten_comma(unique(gene_NMD))) %>%
  mutate(gene_NMD = ifelse(str_detect(gene_NMD, "5pct"),"5pct", ifelse(str_detect(gene_NMD, "10pct"), "10pct","TDP-43 KD only")))

## Compare?
df <- left_join(df_ces,
                df_deps %>% dplyr::select(c(ensgene, (contains("prot")))), 
                by = join_by("ensembl_id" == "ensgene"))

df %>% mutate(lab = ifelse(abs(log2FC_prot) > 0.5, gene_symbol, "")) %>% 
  ggplot(aes(x = psi_max, y = log2FC_prot)) +
  geom_point(alpha=0.8,size=5, aes(fill = factor(gene_NMD)), shape = 21, color = "#363636") + 
  theme_classic() + 
  # gghighlight(gene_NMD != "TDP-43 KD only",unhighlighted_params = list(alpha=0.8, size=5, shape = 16))+
  # ggrepel::geom_text_repel(aes(label = lab), min.segment.length = 1)+
  scale_fill_manual(values=c("#69140E","#e5a4cb", "#BFBFBF")) +
  theme(text = element_text(family="Helvetica")) 


#PATIENTS?
df_deps_pm <- readxl::read_xlsx('/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Projects/NMD/NotGitHub/data/Kozareva_proteomics.xlsx', sheet=15)
df_pm <- left_join(df_ces,
                df_deps_pm %>% dplyr::select(c(external_gene_name, (contains("prot")))) %>% dplyr::select(!c(contains("t__prot"), starts_with("pval__prot"))), 
                by = join_by("gene_symbol" == "external_gene_name")) %>%
  pivot_longer(cols = c(starts_with("log2fc__prot__")), values_to = "log2fc", names_prefix = "log2fc__prot__",  names_to = "dataset") %>%
  pivot_longer(cols = c(starts_with("adjpval__prot__")), values_to = "padj", names_prefix = "adjpval__prot__",  names_to = "dataset_p") %>%
  filter(dataset == dataset_p, !is.na(log2fc), str_detect(dataset,"Guise"), dataset!= "Guise_ALS_MNs_non") %>% dplyr::select(-dataset_p) 

df_pm %>% 
  mutate(dataset = factor(dataset, c("Guise_ALS_MNs_mld", "Guise_ALS_MNs_mod", "Guise_ALS_MNs_sev")),
         log2fc = ifelse(log2fc > 2, 2, ifelse(log2fc < -2, -2, log2fc))) %>%
  ggplot(aes(x = log2fc, y = -log10(padj))) +
  geom_point(alpha=0.8,size=5, aes(fill = factor(gene_NMD)), shape = 21, color = "#363636") + 
  theme_classic() + 
  ggrepel::geom_text_repel(aes(label = gene_symbol), min.segment.length = 0, size = 7, max.overlaps = 5)+
  scale_fill_manual(values=c("#69140E","#e5a4cb", "#BFBFBF")) +
  theme(text = element_text(family="Helvetica"), legend.position = "none",
        panel.spacing = unit(1, "cm", data = NULL)) +
  facet_wrap(vars(dataset), scales="fixed",nrow =1)
  

```
